### 2025-05-27 paper/paper.qmd
Committed paper (almost finalized), references, etc.
Added proper citations via bibtex, updated references, committed paper (will input last bit of writing tomorrow morning).

---
title: "Neighbourhood Crime in Toronto: A Socioeconomic Perspective"
subtitle: "Analyzing Crime Trends in Low, Medium, and High Opportunity Areas (2019–2024)"
author: 
  - name: "Mike Cowan"
    affiliation: "University of Toronto"
thanks: |
  Code and data are available at: [https://github.com/mcowan38/tswd_toronto_crime].
date: "19 May 2025"
date-format: long
execute:
  echo: false
abstract: |
    This study examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 159 Toronto neighbourhoods. Drawing upon Canadian census and Toronto Police crime data, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (High-, Medium-, and Low- Opportunity) and compared crime rates across these clusters over time. Policy implications include targeted interventions in low-opportunity neighbourhoods and continued support for equity-driven community policing and economic opportunity programs.
format: 
  pdf: 
    fig-pos: "H"
    number-sections: true
bibliography: references.bib
---
```{python}
# | label: global_options
# | include: false
# | warning: false
# | message: false
# | echo: false

#### Modules ####
import polars as pl
import numpy as np
import matplotlib.pyplot as plt
from pathlib import Path
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA
from sklearn.cluster import KMeans
from sklearn.mixture import GaussianMixture
from sklearn.metrics import (
    silhouette_score,
    davies_bouldin_score,
    calinski_harabasz_score,
)

#### Display options ####
# Hide column types (i.e., no i64, str, f64 identifiers)
pl.Config.set_tbl_hide_column_data_types(True)

# Force max characters on tables wrap
pl.Config.set_tbl_width_chars(5)

#### Load data ####
crime_raw = pl.read_csv("../data/01-raw_data/neighbourhood_crime.csv")
profile_raw = pl.read_excel("../data/01-raw_data/neighbourhood_profiles.xlsx")
merged_data = pl.read_csv("../data/02-analysis_data/merged_data.csv")

#### Crime setup #####
crime_types = ["assault", "breakenter", "robbery", "shooting"]
years = [2019, 2020, 2021, 2022, 2023, 2024]

#### SES columns #####
ses_columns = [
    "education_rate",
    "prop_single_parent",
    "unemployment_rate",
    "median_income",
]

#### Scaling ####
ses_columns = [
    "education_rate",
    "prop_single_parent",
    "unemployment_rate",
    "median_income",
]
X = merged_data.select(ses_columns).to_numpy()
X_scaled = StandardScaler().fit_transform(X)

#### PCA Coordinates ####
pca_coords = PCA(n_components=2).fit_transform(X_scaled)

#### Configure Clusters ####
cluster_configs = [
    ("KMeans", 2, "Paired", "KMeans (k=2)"),
    ("GMM", 2, "Accent", "GMM (k=2)"),
    ("KMeans", 3, "Set1", "KMeans (k=3)"),
    ("GMM", 3, "Set2", "GMM (k=3)"),
]
```

# Introduction

The COVID-19 pandemic profoundly disrupted social and economic systems worldwide—and crime trends were no exception. In Canada, police-reported incidents plummeted in 2020 before rebounding sharply in 2021 (@statisticscanada_2021_policereported), with violent crime surging 5% above pre-pandemic levels while property crime fell to historic lows.

However, these national trends obscure local disparities; crime clusters unevenly across urban spaces, concentrated in areas marked by structural inequities—e.g., income inequality, educational disparities, and labour market marginalization ([@mohammadi_2022_homicide; @wang_2019_the]). This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods diverge sharply in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage ([@jargowsky_2015_concentrated]). A plethora of research confirms that these inequalities stratify crime patterns, making city-wide or national averages poor proxies for local realities ([@wamg_2019_the; @uesugi_2024; @yu_2025_how]).

Building on this established research, this paper examines how Toronto’s neighbourhood-level crime patterns evolved during the pandemic and its aftermath. We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities—often termed “opportunity” or “advantage” differences ([@yu_2025_how])—relate to diverging crime trends. Specifically, we classify Toronto neighbourhoods into Low-, Medium-, and High-Opportunity clusters via $K$-means clustering on socioeconomic variables and compare the crime trajectories of these clusters over time.

This paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends. Next, we describe our data and methodological approach, justifying our construction of neighbourhood opportunity clusters via $K$-means socioeconomic variables—analyzing their distinct crime trajectories from 2019-2024. Practically, we consider what our neighbourhood-level results suggest for urban policy—particularly targeting policing resources and economic opportunity programs. Finally, we identify directions for future research on urban crime patterns during periods of social disruption.

# Literature Review

## Criminological Theory {#sec-theory}

Criminological theory and empirical evidence converge on two insights in urban crime literature: (1) neighbourhood-level structural conditions fundamentally shape crime, and (2) spatial inequalities between neighbourhoods amplify these effects.

Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls; the resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship and institutional neglect [@frevel_2021_local; @antunes_2021_social]. Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies [@antunes_2021_social], particularly during systemic crises—such as the COVID-19 pandemic.

Contemporary research extends these theories by emphasizing the study of spatial inequality—the juxtaposition of affluence and deprivation across proximate neighbourhoods. For example, Kang (2016) demonstrates that inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence, suggesting that proximity to wealth exacerbates perceptions of exclusion and strain. This finding aligns with @yu_2025_how findings in Paterson, NJ, where median household income exerted the strongest neighbourhood-level influence on crime patterns, underscoring the need to analyze socioeconomic factors at granular spatial scales.

Unsurprisingly, decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@jargowsky_2015_concentrated; @frevel_2021_local]. Toronto exemplifies this dynamic: neighbourhoods with higher marginalization indices exhibit disproportionately high rates of violent and property crime [@wang_2019_the], while areas with entrenched poverty report elevated homicide incidence [@mohammadi_2022_homicide]. However, spatial heterogeneity of these relationships is also critical. Geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods, calling for local (rather than purely global) models of crime.

The COVID-19 pandemic undoubtedly exposed and amplified these spatial inequities. Initial lockdowns reduced city-wide crime through diminished routine activities, but declines in some cities were uneven [@andresen_2022_in]. For instance, while property crime in Vancouver decreased in wealthy neighbourhoods with robust security infrastructure, violent crime surged in disadvantaged areas strained by disrupted social services and weakened guardianship.

These disparities underscore three critical lessons: first, structural disadvantage predicts vulnerability to crime spikes during crises, aligning with social disorganization theory’s emphasis on resource-deprived communities. Second, inter-neighbourhood inequality magnifies criminogenic risks, as strain theory posits when relative deprivation fuels frustration. Third, opportunity structures—access to security, institutional support, and economic stability—are spatially stratified, privileging affluent areas with systemic advantages.

Building on these insights, we operationalize “opportunity” as the structured availability of socioeconomic resources that mediate crime risks, focusing on Toronto’s neighbourhoods. Our analysis extends prior work in two ways: first, we expand the temporal scope to 2019–2024, assessing whether pandemic-era disparities persisted into the recovery phase. Second, we apply $K$-means clustering to classify neighbourhoods into High-, Medium-, and Low-Opportunity clusters using variables (income, education, employment) empirically shown to influence crime patterns [@yu_2025_how]. If social disorganization and strain theories hold, Low-Opportunity clusters should exhibit both higher baseline crime rates and greater volatility during disruptions, reflecting their structural precarity.

# Data {#sec-data}

We use the programming language Python [@python] alongside polars [@polars] and sci-kit-learn [@scikit-learn] for all subsequent data cleaning and analysis.

### Neighbourhood Crime Data

We sourced crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, break-and-enters, robberies, and shootings—the types of crime we chose to analyze between 2019 and 2024. Each entry includes incident years and total counts by neighbourhood (see Table 1).

```{python}
# | label: tbl-crime_data_preview
# | tbl-cap: “Toronto Crime Data (2014-2024)”
# | tbl-cap-location: top
# | echo: false

#### Glimpse crime table ####
crime_glimpse = crime_raw.select(
    [
        # Remove quotes from AREA_NAME values
        pl.col("_id"),
        pl.col("AREA_NAME").str.slice(0, 16),  # keep only first 16 characters
        # Replace nulls with 0 in numeric columns
        pl.col("HOOD_ID"),
        pl.col("ASSAULT_2014").fill_null(0),
        pl.col("BREAKENTER_2014").fill_null(0),
        pl.lit("").alias("Robbery_2014").alias("…"),  # fake truncate
    ]
)

crime_glimpse
```

@tbl-crime_data_preview Previews crime variables across Toronto neighbourhoods, including assault and break-and-enter incidents. Values are cleaned and truncated for display.

### Neighbourhood Census Profiles
2021 Toronto Census Profile data was obtained from the Open Data Toronto Portal [@citeOpenDataToronto]. This data was transformed to align with neighbourhood identifiers used in the crime dataset (see Table 2).

```{python}
# | label: tbl-census_data_preview
# | tbl-cap: “Toronto Neighbourhood Census Data (2021)”
# | tbl-cap-location: top
# | echo: false

#### Glimpse profile table ####
profile_glimpse = (
    profile_raw.select(
        [
            pl.col("Neighbourhood Name").str.slice(0, 24),
            pl.col("West Humber-Clairville").alias("West Humber"),
            pl.col("Mount Olive-Silverstone-Jamestown").alias("Mt. Olive"),
            pl.col("Thistletown-Beaumond Heights").alias("Thistletown"),
            pl.col("Elms-Old Rexdale").alias("Elms-Old"),
            pl.lit("").alias("…"),  # fake truncate
        ]
    )
    .with_row_index("row_nr")
    .filter(pl.col("row_nr") != 1)
    .drop("row_nr")
)

profile_glimpse
```

@tbl-census_data_preview Census-derived socioeconomic variables (e.g., education, income) by neighbourhood, used as clustering inputs. Values are cleaned and truncated for display.

## Measurement

Transforming the lived realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for every neighbourhood and (2) aggregating police‐reported crime counts to rate units.

From the Census data, we extracted four attributes that decades of criminological research identify as correlates of neighbourhood crime: median household income, share of adults with a bachelor’s degree or higher, unemployment rate, and proportion of single-parent families (cite).

From the crime dataset, we aggregated annual counts of assaults, break-and-enters, robberies and shootings (2019–2024) and transformed them into rates per 100,000 persons using neighbourhood-level population denominators.

These two sources were paired by Toronto’s 159 neighbourhoods, yielding socioeconomic features and crime rates from the early pre- to post-COVID-19 pandemic.

# Model

Mirroring prior spatial criminological studies—which have employed diverse spatial statistics to characterize neighbourhood heterogeneity in crime and socio-economic conditions [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how]—we apply $K$-means clustering to the four standardized indicators—education rate, single-parent households, unemployment rate, and median household income—to uncover latent socio-economic typologies.

Let $\mathbf{x}_i = (x_{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, $i = 1, \dots, N = 159$. The features are defined as:

- $x_{i1}$: education rate (proportion with bachelor’s degree or higher)  
- $x_{i2}$: proportion of single-parent households  
- $x_{i3}$: unemployment rate  
- $x_{i4}$: median household income

We standardize each raw feature via  
$$
x_{ij}^{\mathrm{scaled}} = \frac{x_{ij} - \overline{x}_j}{s_j},
$$  
where $\overline{x}_j$ and $s_j$ are the sample mean and standard deviation of feature $j$.

We then partition $\{\mathbf{x}_i\}_{i=1}^N$ into $K = 3$ clusters using the $K$-means algorithm, minimizing the within-cluster sum of squares:  
$$
\{\boldsymbol{\mu}_k\}_{k=1}^3
= \arg\min_{\{\mu_k\}}
\sum_{i=1}^N \bigl\|\mathbf{x}_i - \mu_{z_i}\bigr\|^2,
$$  
where $\mu_{z_i}$ is the centroid of neighbourhood $i$’s assigned cluster.

We tested $K = 2, \dots, 5$ and selected $K = 3$ based on silhouette and Davies–Bouldin diagnostics (see Appendix{sec-}), and because a three-fold typology (High-, Medium-, Low-Opportunity) aligns with established frameworks of urban socioeconomic stratification. A preliminary Principal Components Analysis—conducted as a diagnostic check—showed that the first two components explain ~89% of total variance, confirming a strong low-dimensional socioeconomic gradient underlying the four indicators.

The resulting clusters are labeled as:
```{python}
# | label: tbl-cluster_stats
# | tbl-cap: “Descriptive Statistics by SES Cluster (2019–2024)”
# | tbl-cap-location: top
# | echo: false

#### Descriptive Statistics ####
# Summary by cluster
cluster_summary = (
    merged_data.group_by("cluster")
    .agg(
        [
            pl.count("neighbourhood").alias("Neighbourhoods (n)"),
            pl.mean("median_income").alias("Avg. Median Income ($)"),
            pl.mean("prop_single_parent").round(2).alias("Single-Parent Proportion"),
            (pl.mean("education_rate")).round(2).alias("Education Rate"),
            (pl.mean("unemployment_rate")).round(2).alias("Unemployment Rate"),
        ]
    )
    .sort("cluster")
    .rename({"cluster": "Cluster"})
)

#### Display Quarto table ####
cluster_summary
```

@tbl-cluster_stats Presents the average socioeconomic features by cluster for Toronto neighbourhoods over 2019–2024. Clusters were derived via $K$-means clustering on standardized education rate, proportion of single-parent households, unemployment rate, and median household income. High-Opportunity (Cluster 0) neighbourhoods exhibit high educational attainment and median income alongside low unemployment and single-parent household rates. Medium-Opportunity (Cluster 1) neighbourhoods display a mixed socioeconomic profile. Low-Opportunity (Cluster 2) neighbourhoods are characterized by lower education and income, and higher unemployment and single-parent household rates. All values represent means of each indicator across the study period.

Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating for each component a mean vector and covariance matrix  both methods yielded similar groupings [@scikit-learn]. We retained $K$-means for its interpretability and marginally better cluster evaluation scores (see @sec-appendix)

# Results

Our results are summarized in @fig-crime_trends

```{python}
#| label: fig-crime_trends
#| fig-cap: "Average Crime Rate Trends (2019–2024) by Socio-Economic Cluster"
#| fig-subcap: ["Assault", "Break-and-Enter", "Robbery", "Shooting"]
#| echo: false

### Trends plot ####
# Create combined 2x2 figure
fig, axes = plt.subplots(2, 2, figsize=(12, 8))
axes = axes.flatten()

# Loop through each crime type and plot trends by SES cluster #
for idx, crime in enumerate(crime_types):
    rate_cols = [f"{crime}_rate_{year}" for year in years]

    # Unpivot wide to long for time-series, extract year as integer
    merged_long = (
        merged_data.select(["Socioeconomic_index", *rate_cols])
        .unpivot(
            on=rate_cols,
            index="Socioeconomic_index",
            variable_name="year",
            value_name="rate",
        )
        .with_columns(pl.col("year").str.extract(r"_(\d{4})$", 1).cast(pl.Int64))
    )

    # Group by SES and year
    trend = (
        merged_long.group_by(["Socioeconomic_index", "year"])
        .agg(pl.col("rate").mean().alias("average_rate"))
        .sort(["Socioeconomic_index", "year"])
    )

    # Plot in the 2x2 subplot
    ax = axes[idx]
    for group in ["Low Opportunity", "Medium Opportunity", "High Opportunity"]:
        sub = trend.filter(pl.col("Socioeconomic_index") == group)
        ax.plot(
            sub["year"].to_list(),
            sub["average_rate"].to_list(),
            marker="o",
            label=group,
        )

    ax.set_title(f"{crime.title()} Rate")
    ax.set_xlabel("Year")
    ax.set_ylabel("Rate per 100K persons")
    ax.legend(title="SES Cluster", fontsize=8)

#### Display ####
fig.suptitle("Crime Rate Trends by Socio-Economic Cluster (2019–2024)", fontsize=16)
fig.tight_layout(rect=[0, 0.03, 1, 0.95])
plt.show()
```

@fig-crime_trends Plots...

# Discussion

## Neighbourhood Crime Rates

- ETA: May 28

## Theoretical Implications

- ETA: May 28

## Policy Relevance
- ETA: May 28

## Limitations {#sec-limitations}

- ETA: May 28

\newpage

\appendix

# Appendix {#sec-appendix}

## Model Evaluation {#sec-model-evaluation}

- ETA: May 28

```{python}
# | label: fig-cluster_pca
# | fig-cap: "Dimensionality‐Reduced Clustering Comparisons (2019–2024 SES)"
# | fig-subcap: ["KMeans (k=2)", "GMM (k=2)", "KMeans (k=3)", "GMM (k=3)"]
# | fig-width: 6.5
# | fig-height: 5
# | echo: false

#### Model Evaluation ####
fig, axes = plt.subplots(2, 2, figsize=(12, 10))
axes = axes.flatten()

for i, (model_type, k, cmap, title) in enumerate(cluster_configs):
    # instantiate
    if model_type == "KMeans":
        model = KMeans(n_clusters=k, random_state=42, n_init=10)
    else:
        model = GaussianMixture(n_components=k, random_state=42)
    labels = model.fit_predict(X_scaled)

    # Scatterplot
    sc = axes[i].scatter(
        pca_coords[:, 0], pca_coords[:, 1], c=labels, cmap=cmap, alpha=0.8
    )
    axes[i].set_title(title)
    axes[i].set_xlabel("PC 1")
    axes[i].set_ylabel("PC 2")
    axes[i].legend(*sc.legend_elements(), title="Cluster")

fig.suptitle("PCA‐Based Clustering: KMeans vs. GMM", fontsize=16)
fig.tight_layout(rect=[0, 0.03, 1, 0.95])
plt.show()
```

@fig-cluster_pca Displays...

Further, [ETA: May 28]
```{python}
# | label: fig-cluster_metrics
# | fig-cap: "Clustering Evaluation Metrics by Model and k"
# | fig-subcap: ["Silhouette Score", "Davies–Bouldin Index", "Calinski–Harabasz Score"]
# | fig-width: 6.5
# | fig-height: 5
# | echo: false

#### Compute Evaluation Metrics ####
evals = []
for model_type in ["KMeans", "GMM"]:
    for k in [2, 3]:
        if model_type == "KMeans":
            m = KMeans(n_clusters=k, random_state=42, n_init=10)
        else:
            m = GaussianMixture(n_components=k, random_state=42)
        lbls = m.fit_predict(X_scaled)
        evals.append(
            {
                "Model": model_type,
                "k": k,
                "Silhouette": silhouette_score(X_scaled, lbls),
                "Davies-Bouldin": davies_bouldin_score(X_scaled, lbls),
                "Calinski-Harabasz": calinski_harabasz_score(X_scaled, lbls),
            }
        )

#### Plot ####
metrics = ["Silhouette", "Davies-Bouldin", "Calinski-Harabasz"]
titles = [
    "Silhouette Score (↑ better sep.)",
    "Davies–Bouldin Index (↓ less overlap)",
    "Calinski–Harabasz Score (↑ tighter clusters)",
]

fig, axs = plt.subplots(1, 3, figsize=(18, 5))
for idx, metric in enumerate(metrics):
    ax = axs[idx]
    x = np.arange(2)  # k = 2, 3
    width = 0.35
    for j, model_type in enumerate(["KMeans", "GMM"]):
        scores = [e[metric] for e in evals if e["Model"] == model_type]
        ax.bar(x + (j - 0.5) * width, scores, width, label=model_type)
    ax.set_xticks(x)
    ax.set_xticklabels(["k=2", "k=3"])
    ax.set_title(titles[idx])
    ax.set_xlabel("Clusters (k)")
    ax.set_ylabel(metric)
    ax.legend()

fig.suptitle("Clustering Metrics: KMeans vs. GMM", fontsize=16)
fig.tight_layout(rect=[0, 0.03, 1, 0.95])
plt.show()
```

@fig-cluster_metrics

\newpage

# References

### 2025-05-27 scripts/08-plot_crime_clusters.py

@@ -11,7 +11,7 @@
 #### Workspace setup ####
 import polars as pl
 import matplotlib.pyplot as plt
-from pathlib import Path
+from pathlib import Path  # inherent to Python
 
 #### Load data ####
 merged_data = pl.read_csv("data/02-analysis_data/merged_data.csv")

### 2025-05-27 scripts/09-model_evaluation.py

@@ -82,6 +82,23 @@ fig_pca.suptitle(
 fig_pca.tight_layout(rect=[0, 0.03, 1, 0.95])
 fig_pca.savefig("other/figures/fig_2_cluster_comparisons.png", dpi=300)
 
+#### Check PCA scores ####
+# Fit PCA on the scaled matrix (using all components)
+pca_full = PCA().fit(scaled_matrix)
+
+# Explained variance ratio per component
+print("Explained variance ratio:", pca_full.explained_variance_ratio_)
+
+# Cumulative explained variance
+print("Cumulative explained variance:", pca_full.explained_variance_ratio_.cumsum())
+
+# Singular values (proportional to component strengths)
+print("Singular values:", pca_full.singular_values_)
+
+# Eigenvalues (variance captured by each principal axis)
+eigenvalues = pca_full.explained_variance_
+print("Eigenvalues:", eigenvalues)
+
 #### Cluster Metrics Evaluation ####
 # Compare models using Silhouette, Davies-Bouldin, and Calinski-Harabasz scores
 # [https://scikit-learn.org/stable/modules/clustering.html#clustering-evaluation]

### 2025-05-28 paper/paper.qmd
Pushed Final draft update.

@@
 ---
-title: "Neighbourhood Crime in Toronto: A Socioeconomic Perspective"
-subtitle: "Analyzing Crime Trends in Low, Medium, and High Opportunity Areas (2019–2024)"
+title: "Neighbourhood Crime in Toronto (2019–2024)"
+subtitle: "Analyzing Crime Trends in Low, Medium, and High Socioeconomic Areas"
 author: 
@@
   echo: false
-abstract: |
-    This study examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 159 Toronto neighbourhoods.
-Drawing upon Canadian census and Toronto Police crime data, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (High-, Medium-, and Low- Opportunity) and compared crime rates across these clusters over time.
-Policy implications include targeted interventions in low-opportunity neighbourhoods and continued support for equity-driven community policing and economic opportunity programs.
+abstract: This paper examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 159 Toronto neighbourhoods.
+Drawing on data from the 2021 Canadian Census and the Toronto Police Service, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (High-, Medium-, and Low- Opportunity) and compared crime rates across these clusters over time.
+Across all four crime types, Low Opportunity neighbourhoods experienced persistently higher rates—especially for assault and shootings—while High Opportunity areas remained relatively insulated.
+Following a city-wide decline in break-and-enters during the 2020 lockdown, crime rebounded steeply in deprived areas, with robbery and gun violence disproportionately escalating through 2023.
+These findings highlight persistent socioeconomic disparities in crime exposure, with disadvantaged neighbourhoods bearing a disproportionate share of urban crime.
 format: 
@@
     bibliography: references.bib
+    csl: styles/apa.csl
+
 ---
@@
 import matplotlib.pyplot as plt
+import matplotlib as mpl
+import matplotlib.font_manager as fm
 from pathlib import Path
@@
 
-#### Display options ####
+#### Standardize Figure Font in matplotlib ####
+# [https://matplotlib.org/stable/users/explain/text/usetex.html]
+# [https://ctan.org/pkg/cm-unicode]
+
+font_path = Path("fonts/cmunrm.otf").resolve()
+fm.fontManager.addfont(font_path)
+cmu_font = fm.FontProperties(fname=str(font_path))
+
+mpl.rcParams.update(
+    {
+        "font.family": cmu_font.get_name(),  # Should be "CMU Roman"
+        "mathtext.fontset": "cm",
+        "axes.unicode_minus": False,
+    }
+)
+
 # Hide column types (i.e., no i64, str, f64 identifiers)
@@
 profile_raw = pl.read_excel("../data/01-raw_data/neighbourhood_profiles.xlsx")
-merged_data = pl.read_csv("../data/02-analysis_data/merged_data.csv")
+merged_data = pl.read_csv("../data/02-analysis_data/02-analysis_data_merged.csv")
+rates_change = pl.read_csv("../data/02-analysis_data/04-cluster_crime_rates.csv")
 
@@
 
-The COVID-19 pandemic profoundly disrupted social and economic systems worldwide—and crime trends were no exception.
-In Canada, police-reported incidents plummeted in 2020 before rebounding sharply in 2021 (@statisticscanada_2021_policereported), with violent crime surging 5% above pre-pandemic levels while property crime fell to historic lows.
+The COVID-19 pandemic disrupted social and economic systems worldwide—and crime trends were no exception.
+In Canada, police-reported incidents plummeted in 2020 before rebounding sharply in 2021 (Statistics Canada, 2021), with violent crime surging 5% above pre-pandemic levels while property crime fell to historic lows.
 
-However, these national trends obscure local disparities; crime clusters unevenly across urban spaces, concentrated in areas marked by structural inequities—e.g., income inequality, educational disparities, and labour market marginalization ([@mohammadi_2022_homicide; @wang_2019_the]).
-This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods diverge sharply in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage ([@jargowsky_2015_concentrated]).
-A plethora of research confirms that these inequalities stratify crime patterns, making city-wide or national averages poor proxies for local realities ([@wamg_2019_the; @uesugi_2024; @yu_2025_how]).
+However, these national trends obscure local disparities; crime clusters unevenly across urban spaces, concentrated in areas marked by structural inequities—e.g., income inequality, educational disparities, and labour market marginalization [@mohammadi_2022_homicide; @wang_2019_the].
+This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods diverge sharply in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage [@jargowsky_2015_concentrated].
+A plethora of research confirms that these inequalities stratify crime patterns, making city-wide or national averages poor proxies for local realities [@wang_2019_the; @uesugi_2024_a; @yu_2025_how].
 
 Building on this established research, this paper examines how Toronto’s neighbourhood-level crime patterns evolved during the pandemic and its aftermath.
-We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities—often termed “opportunity” or “advantage” differences ([@yu_2025_how])—relate to diverging crime trends.
-Specifically, we classify Toronto neighbourhoods into Low-, Medium-, and High-Opportunity clusters via $K$-means clustering on socioeconomic variables and compare the crime trajectories of these clusters over time.
+We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities—often termed “opportunity” or “advantage” differences [@yu_2025_how]—relate to diverging crime trends.
+Specifically, we classify Toronto neighbourhoods into Low-, Medium-, and High-Opportunity clusters via $K$-means clustering on socioeconomic proxies and compare the crime trajectories of these clusters over time.
 
 This paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends.
-Next, we describe our data and methodological approach, justifying our construction of neighbourhood opportunity clusters via $K$-means socioeconomic variables—analyzing their distinct crime trajectories from 2019-2024.
-Practically, we consider what our neighbourhood-level results suggest for urban policy—particularly targeting policing resources and economic opportunity programs.
-Finally, we identify directions for future research on urban crime patterns during periods of social disruption.
+Next, we describe our data and methodological approach, justifying our construction of neighbourhood opportunity clusters and analyzing their distinct crime trajectories from 2019-2024.
+Practically, we consider what our neighbourhood-level results suggest for urban crime policy.
 
@@
 
-## Criminological Theory {#sec-theory}
-
 Criminological theory and empirical evidence converge on two insights in urban crime literature: (1) neighbourhood-level structural conditions fundamentally shape crime, and (2) spatial inequalities between neighbourhoods amplify these effects.
 
-Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls; the resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship and institutional neglect [@frevel_2021_local; @antunes_2021_social].
-Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies [@antunes_2021_social], particularly during systemic crises—such as the COVID-19 pandemic.
+Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls.
+The resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship and institutional neglect [@antunes_2021_social; @frevel_2021_local].
+Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies particularly during systemic crises-e.g., the COVID-19 pandemic [@antunes_2021_social].
 
@@
 For example, Kang (2016) demonstrates that inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence, suggesting that proximity to wealth exacerbates perceptions of exclusion and strain.
-This finding aligns with @yu_2025_how findings in Paterson, NJ, where median household income exerted the strongest neighbourhood-level influence on crime patterns, underscoring the need to analyze socioeconomic factors at granular spatial scales.
+These findings align with contemporary research in Paterson, NJ, which suggested median household income exerted the strongest neighbourhood-level influence on crime patterns [@yu_2025_how's].
 
-Unsurprisingly, decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@jargowsky_2015_concentrated; @frevel_2021_local].
+Unsurprisingly, decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@frevel_2021_local; @jargowsky_2015_concentrated].
 Toronto exemplifies this dynamic: neighbourhoods with higher marginalization indices exhibit disproportionately high rates of violent and property crime [@wang_2019_the], while areas with entrenched poverty report elevated homicide incidence [@mohammadi_2022_homicide].
-However, spatial heterogeneity of these relationships is also critical.
-Geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods, calling for local (rather than purely global) models of crime.
+However, the spatial heterogeneity of these relationships in major cities is also critical.
+For example, geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods, demonstrating the need for crime models that are sensitive to local socioeconomic dynamics.
 
@@
 Second, inter-neighbourhood inequality magnifies criminogenic risks, as strain theory posits when relative deprivation fuels frustration.
-Third, opportunity structures—access to security, institutional support, and economic stability—are spatially stratified, privileging affluent areas with systemic advantages.
+Third, opportunity structures—access to security, institutional support, and economic stability—can be spatially stratified, privileging affluent areas with systemic advantages.
 
-Building on these insights, we operationalize “opportunity” as the structured availability of socioeconomic resources that mediate crime risks, focusing on Toronto’s neighbourhoods.
-Our analysis extends prior work in two ways: first, we expand the temporal scope to 2019–2024, assessing whether pandemic-era disparities persisted into the recovery phase.
+Building on these insights, we operationalize “opportunity” as relative access to stabilizing socioeconomic conditions—a small composite measure proxied by median income, educational attainment, employment status, and household structure.
+Our analysis, focusing on Toronto neighbourhoods, compliments prior research in two ways: first, we expand the temporal scope to 2019–2024, assessing whether pandemic-era disparities persisted into the recovery phase.
 Second, we apply $K$-means clustering to classify neighbourhoods into High-, Medium-, and Low-Opportunity clusters using variables (income, education, employment) empirically shown to influence crime patterns [@yu_2025_how].
@@
 
-We use the programming language Python [@python] alongside polars [@polars] and sci-kit-learn [@scikit-learn] for all subsequent data cleaning and analysis.
+We use the programming language Python [@python] alongside polars [@polars] and sci-kit-learn [@scikit-learn] for the paper's data cleaning and analysis.\footnote{Workflow and GitHub-based project structure follow best practices guided by Alexander (2023), which has proven to be an excellent resource: [https://tellingstorieswithdata.com]}.
 
@@
 We sourced crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, break-and-enters, robberies, and shootings—the types of crime we chose to analyze between 2019 and 2024.
-Each entry includes incident years and total counts by neighbourhood (see Table 1).
+Each entry includes incident years and total counts by neighbourhood (see @tbl-crime_data_preview).
 
@@
 # | label: tbl-crime_data_preview
-# | tbl-cap: “Toronto Crime Data (2014-2024)”
+# | tbl-cap: Toronto Crime Data (2014-2024)
 # | tbl-cap-location: top
@@
 
+#### Display ####
 crime_glimpse
@@
 
-@tbl-crime_data_preview Previews crime variables across Toronto neighbourhoods, including assault and break-and-enter incidents. Values are cleaned and truncated for display.
+*Note.* @tbl-crime_data_preview previews Toronto Police crime data for 2014–2024. Columns include neighbourhood identifiers and annual counts of major crimes, truncated for display.
 
@@
 2021 Toronto Census Profile data was obtained from the Open Data Toronto Portal [@citeOpenDataToronto].
-This data was transformed to align with neighbourhood identifiers used in the crime dataset (see Table 2).
+This data was transformed to align with neighbourhood identifiers used in the crime dataset (see @tbl-census_data_preview).
 
@@
 # | label: tbl-census_data_preview
-# | tbl-cap: “Toronto Neighbourhood Census Data (2021)”
+# | tbl-cap: Toronto Neighbourhood Census Data (2021)
 # | tbl-cap-location: top
@@
 
+#### Display ####
 profile_glimpse
@@
 
-@tbl-census_data_preview Census-derived socioeconomic variables (e.g., education, income) by neighbourhood, used as clustering inputs. Values are cleaned and truncated for display.
+*Note.* @tbl-census_data_preview previews the 2021 Toronto Census dataset.
+Columns include standardized socioeconomic indicators (e.g., education rate, household income) across selected neighbourhoods.
+Names and values have been truncated for display.
 
@@
 
-Transforming the lived realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for every neighbourhood and (2) aggregating police‐reported crime counts to rate units.
+Transforming the lived realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for each neighbourhood and (2) aggregating police‐reported crime counts to rate units.
 
-From the Census data, we extracted four attributes that decades of criminological research identify as correlates of neighbourhood crime: median household income, share of adults with a bachelor’s degree or higher, unemployment rate, and proportion of single-parent families (cite).
+From the Census data, we extracted four attributes that decades of criminological research identify as correlates of neighbourhood crime: median household income, share of adults with a bachelor’s degree or higher, unemployment rate, and proportion of single-parent families.
 
@@
 
-Mirroring prior spatial criminological studies—which have employed diverse spatial statistics to characterize neighbourhood heterogeneity in crime and socio-economic conditions [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how]—we apply $K$-means clustering to the four standardized indicators—education rate, single-parent households, unemployment rate, and median household income—to uncover latent socio-economic typologies.
+Cognizant of prior criminological studies—which have employed powerful spatial models to characterize neighbourhood heterogeneity in crime and socioeconomic conditions (utilizing substantially more predictors; [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how])—we apply $K$-means clustering to four standardized indicators—education rate, proportion of single-parent households, unemployment rate, and median household income—to cluster neighbourhoods based on their socioeconomic similarity.
 
-Let $\mathbf{x}_i = (x_{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, $i = 1, \dots, N = 159$. The features are defined as:
+The $K$-means algorithm identifies the set of cluster centroids ${\mu_k}$ that minimizes the total within-cluster sum of squared distances (WCSS)\footnote{Mathematicals have been adapted from Hastie et al.
+(2009).}.
+In simple terms, the algorithm partitions observations into $K$ groups such that data points within each group are as similar as possible to one another.
+Similarity is measured by the squared Euclidean distance between each observation and the average of its assigned group—called the cluster centroid.
+The goal is to find centroid locations that minimize the total variation within each cluster, producing compact and well-separated groups.
 
-- $x_{i1}$: education rate (proportion with bachelor’s degree or higher)  
-- $x_{i2}$: proportion of single-parent households  
-- $x_{i3}$: unemployment rate  
-- $x_{i4}$: median household income
+Let $\mathbf{x}i = (x{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, where $i = 1, \dots, N = 159$.
+Each vector includes standardized values for (1) median household income, (2) the proportion of residents with a bachelor’s degree or higher, (3) the unemployment rate, and (4) the single-parent household rate.
 
-We standardize each raw feature via  
+Each raw feature os standardized by 
 $$
@@
 $$
+
 where $\overline{x}_j$ and $s_j$ are the sample mean and standard deviation of feature $j$.
 
-We then partition $\{\mathbf{x}_i\}_{i=1}^N$ into $K = 3$ clusters using the $K$-means algorithm, minimizing the within-cluster sum of squares:  
+We then partition $\{\mathbf{x}_i\}_{i=1}^N$ into $K = 3$ clusters using the $K$-means algorithm, minimizing the WCSS:  
 $$
@@
 $$
+
 where $\mu_{z_i}$ is the centroid of neighbourhood $i$’s assigned cluster.
 
-We tested $K = 2, \dots, 5$ and selected $K = 3$ based on silhouette and Davies–Bouldin diagnostics (see Appendix{sec-}), and because a three-fold typology (High-, Medium-, Low-Opportunity) aligns with established frameworks of urban socioeconomic stratification.
-A preliminary Principal Components Analysis—conducted as a diagnostic check—showed that the first two components explain ~89% of total variance, confirming a strong low-dimensional socioeconomic gradient underlying the four indicators.
+We tested $K = 2, \dots, 5$ and selected $K = 3$ based on silhouette diagnostics (see @sec-model-evaluation).
+Additionally, Principal Components Analysis—conducted as a diagnostic check—showed that the first two components explain ~89% of total variance, confirming a strong low-dimensional socioeconomic gradient underlying the four indicators.
 
-The resulting clusters are labeled as:
+We adopted a three-group classification—High, Medium, and Low Opportunity—aligned with standard socioeconomic strata commonly used in policy discourse and everyday language (i.e., upper, middle, and lower class).
+Based on this scheme, the resulting clusters are labelled as shown in @tbl-cluster_stats below.
 ```{python}
 # | label: tbl-cluster_stats
-# | tbl-cap: “Descriptive Statistics by SES Cluster (2019–2024)”
+# | tbl-cap: Descriptive Statistics by SES Cluster (2019–2024)
 # | tbl-cap-location: top
@@
 
-#### Display Quarto table ####
+#### Display ####
 cluster_summary
@@
 
-@tbl-cluster_stats Presents the average socioeconomic features by cluster for Toronto neighbourhoods over 2019–2024.
+*Note.* @tbl-cluster_stats presents the average socioeconomic features by cluster for Toronto neigbourhoods over 2019–2024.
 Clusters were derived via $K$-means clustering on standardized education rate, proportion of single-parent households, unemployment rate, and median household income.
@@
 All values represent means of each indicator across the study period.
+**Needs reshaping**
 
-Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating for each component a mean vector and covariance matrix  both methods yielded similar groupings [@scikit-learn].
-We retained $K$-means for its interpretability and marginally better cluster evaluation scores (see @sec-appendix)
+Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating a mean vector and covariance matrix for each component.}—which relax the spherical cluster assumption imposed by $K$-means [@scikit-learn—we], we retained the $K$-means technique due to its superior evaluation scores and greater interpretability (both methods produced broadly similar neighbourhood groupings; see @fig-cluster_pca).
 
@@
 
-Our results are summarized in @fig-crime_trends
+Our results are summarized in @fig-crime_trends.
 
 ```{python}
-#| label: fig-crime_trends
-#| fig-cap: "Average Crime Rate Trends (2019–2024) by Socio-Economic Cluster"
-#| fig-subcap: ["Assault", "Break-and-Enter", "Robbery", "Shooting"]
-#| echo: false
+# | label: fig-crime_trends
+# | fig-cap: Average Crime Rate Trends by Socioeconomic Cluster (2019–2024)
+# | fig-cap-location: top
+# | echo: false
 
@@
     merged_long = (
-        merged_data.select(["Socioeconomic_index", *rate_cols])
+        merged_data.select(
+            ["Socioeconomic_index", *rate_cols])
         .unpivot(
@@
 #### Display ####
-fig.suptitle("Crime Rate Trends by Socio-Economic Cluster (2019–2024)", fontsize=16)
 fig.tight_layout(rect=[0, 0.03, 1, 0.95])
@@
 
-@fig-crime_trends Plots...
+*Note.* @fig-crime_trends plots trends in crime rates for Assault, Break-and-Enter, Robbery, and Shootings.
+For each year and crime type, the figure shows the mean rate across all neighbourhoods within each cluster.
+
+Figure @fig-crime_trends illustrates how crime unfolded across Toronto’s neighbourhoods between 2019 and 2024.
+In Low-Opportunity neighbourhoods, assault rates edged upward from 758 to 805 incidents per 100,000 persons—an increase of 6.2%—while Medium-Opportunity areas saw an 8.0% rise, from 737 to 796 per 100,000.
+Even High-Opportunity neighbourhoods experienced a modest 8.0% uptick in assaults, climbing from 446 to 482 per 100,000.
+
+By contrast, property-type and violent crimes more pronounced divergence.
+Robbery rates in High-Opportunity areas plunged by roughly one-third (a 34.5% reduction), falling from 95 to 62 per 100,000, whereas Low-Opportunity clusters recorded a smaller 16.9% decline (126 → 104).
+Medium-Opportunity wards registered an 18.5% drop (124 → 101).
+Shooting incidents fell most dramatically in Low-Opportunity areas—dropping from 27.95 to 15 per 100,000, a 44.0% decrease—while Medium- and High-Opportunity neighbourhoods saw reductions of 19.0% (10 → 8) and 18.5% (5 → 4), respectively.
+
+The 2020 city-wide dip in break-and-enters—driven, in part, by greater residential presence—was particularly deep in deprived areas, where rates fell 26.0% from 246 to 182 per 100,000 between 2019 and 2020.
+However, by 2023 Low-Opportunity wards had rebounded to 214 incidents per 100,000—just 12.9% below their pre-pandemic level.
+Medium-Opportunity neighbourhoods saw a more sustained decline, with break-and-enters down 24.1% (329 → 249), while High-Opportunity areas only edged 2.2% above 2019 levels (348 → 356)\footnote{See Table 3 for tabular breakdown of year-to-year rate change.}.
 
@@
 
-## Neighbourhood Crime Rates
+The patterns we observed across Toronto’s neighbourhoods from 2019 to 2024 align closely with the expectations derived from social disorganization and strain theories.
+As predicted, Low Opportunity areas—characterized by lower income, education, and employment levels, and higher single-parent prevalence—exhibited consistently elevated crime rates across all four crime types.
+Expectedly, these neighbourhoods began the pre-pandemic era with higher baseline crime levels but also demonstrated greater volatility in response to pandemic-related shocks.
 
-- ETA: May 28
+Social disorganization theory helps explain these patterns by highlighting how concentrated disadvantage weakens collective efficacy and informal social control mechanisms [@frevel_2021_local].
+The relatively rapid rebound in property crime and the persistent elevation of violent crime in less advantaged areas suggest that these neighbourhoods may lack the institutional buffers and community cohesion necessary to sustain safety gains during and after the 2020 lockdown.
+In contrast, High Opportunity neighbourhoods—presumably richer in social capital and surveillance infrastructure—remained largely insulated from post-pandemic escalations [@wang_2019_the].
 
-## Theoretical Implications
+Strain theory offers a complementary lens for interpreting the steeper rise in robbery and shootings in Low Opportunity clusters following 2020.
+These forms of violent and acquisitive crime may reflect cumulative stressors—e.g., economic hardship, institutional disengagement, and perceived exclusion—amplified by the pandemic [@antunes_2021_social].
+The fact that Medium Opportunity areas also experienced upward trends in robbery and shootings suggests a broader diffusion of strain-related pressures, albeit less intensely than has been observed in structurally marginalized zones [@yu_2025_how].
 
-- ETA: May 28
+Importantly, the growing gap between Low and Medium clusters—particularly in rates of robbery and gun violence—could indicate a deepening of intra-urban inequality.
+This dynamic resonates with spatial criminological findings that crime concentration is not only persistent but intensifying in the most structurally vulnerable areas [@mohammadi_2022_homicide; @uesugi_2024_a].
 
-## Policy Relevance
-- ETA: May 28
+These findings reinforce multiple insights from prior criminology literature: (1) Structural disadvantage is predictive of elevated and volatile crime exposure; (2) Neighbourhoods vary in their capacity to absorb or rebound from systemic shocks [@andresen_2022_in]; (3) Socioeconomic opportunity is not merely correlated with crime rates, but mediates the resilience of communities during periods of crisis [@jargowsky_2015_concentrated, @kang_2016_inequality].
+
+Taken together, the results underscore the importance of continuing to analyze urban crime through the dual lenses of structural inequality and spatial heterogeneity.
+The typological clustering approach employed here—grounded in census-derived socioeconomic indicators—provides a simple, replicable method for the first-pass identification of potential areas of concern, and understanding how patterns of crime map onto entrenched dimensions of social disorganization and strain.
 
@@
 
-- ETA: May 28
+This study has several data-oriented limitations.
+First, our analysis is ecological in nature, drawing conclusions from neighbourhood-level aggregates rather than individual-level data.
+As such, we cannot determine whether individuals with certain socioeconomic characteristics are more likely to experience or perpetrate crime; we can only observe area-level patterns.
+
+Second, the analysis relies solely on police-reported crime data, which are subject to well-documented limitations.
+Not all crimes are reported to law enforcement, and fluctuations in patrol coverage, resource allocation, or surveillance infrastructure can artificially inflate or deflate recorded incidents.
+These reporting inconsistencies likely vary across neighbourhoods, potentially skewing cross-area comparisons.
+
+Additionally, the socioeconomic data we use are static, drawn from the 2021 Census.
+These indicators—income, education, unemployment, and single-parent status—cannot capture changes that occurred before or after 2021.
+Moreover, we excluded several potentially important variables—such as neighbourhood ethnic composition, immigration status, and residential mobility—which may shape both crime exposure and neighbourhood opportunity in substantive ways.
+As a result, our clustering captures only a partial view of structural disadvantage.
+
+Fourth, our method for grouping neighbourhoods—$K$-means clustering—makes simplifying assumptions.
+It treats all four variables as equally important and assumes the clusters are roughly circular and evenly sized-while in reality, socioeconomic differences are rarely that tidy (@scikit-learn).
+Alternative clustering methods (e.g., GMM) might better reflect the true structure of the data.
+
+Future research should address these gaps by using dynamic, multi-year socioeconomic data, incorporating a broader range of variables, and accounting for spatial relationships between neighbourhoods.
+In addition, combining quantitative patterns with qualitative accounts of neighbourhood residents could also help ground the numbers in the lived experiences of residents.
 
@@
 
-- ETA: May 28
+### Cluster Plots
+A choice of socioeconomic index can meaningfully shift the resulting clusters.
+One straightforward approach is to standardize each indicator (e.g., income, education, unemployment) to z-scores and sum them [@vyas_2006_constructing].
+Alternatively, factor analysis, which derives weights from latent dimensions (e.g., “Economic Capital” or “Family Stability”) can better reflect the structure among variables but requires more complex modelling and larger samples.
+
+Given our inquiry is exploratory, we opted to employ $K$-Means.
+@fig-cluster_pca reduces our data to the first two principal components for visualization purposes.
+To guard against algorithmic artifacts, we place $K$-Means (hard, spherical boundaries) alongside Gaussian Mixture Models (soft, ellipsoidal groupings) [@scikit-learn].
+Consistency across both methods strengthens confidence that our identified neighbourhood clusters reflect real socioeconomic patterns (and that we made the right selection).
 
@@
 # | label: fig-cluster_pca
-# | fig-cap: "Dimensionality‐Reduced Clustering Comparisons (2019–2024 SES)"
-# | fig-subcap: ["KMeans (k=2)", "GMM (k=2)", "KMeans (k=3)", "GMM (k=3)"]
-# | fig-width: 6.5
-# | fig-height: 5
+# | fig-cap: Clustering Techniques Comparison ($k = 2, 3$)
+# | fig-cap-location: top
 # | echo: false
@@
     # Scatterplot
-    sc = axes[i].scatter(
+    scatterplot = axes[i].scatter(
         pca_coords[:, 0], pca_coords[:, 1], c=labels, cmap=cmap, alpha=0.8
@@
     axes[i].set_ylabel("PC 2")
-    axes[i].legend(*sc.legend_elements(), title="Cluster")
+    axes[i].legend(*scatterplot.legend_elements(), title="Cluster")
 
-fig.suptitle("PCA‐Based Clustering: KMeans vs. GMM", fontsize=16)
 fig.tight_layout(rect=[0, 0.03, 1, 0.95])
@@
 
-@fig-cluster_pca Displays...
+*Note.* @fig-cluster_pca displays the two-dimensional projection of neighbourhoods, coloured by assigned cluster.
+At $k = 3$, GMM’s middle component (Cluster 1) overlaps with the Cluster 2, suggesting less clear separation at three clusters.
+Visually, $K$-Means ($k = 3$) appears to be the superior model choice.
 
-Further, [ETA: May 28]
+### Evaluation Metrics
+
+@fig-cluster_metrics presents a detailed comparison of clustering quality across $K$-means and Gaussian mixture models.
+We compute the Silhouette Score (ranging from –1 to 1; higher values indicate better cohesion and separation) to assess how closely each point is matched to its own cluster versus the next best alternative.
+The Davies–Bouldin Index measures the average similarity between each cluster and its most similar counterpart—lower values denote more compact, well-separated clusters.
+Finally, the Calinski–Harabasz Score quantifies the ratio of between-cluster dispersion to within-cluster dispersion—higher scores signify clearer, more distinct grouping.
+
 ```{python}
 # | label: fig-cluster_metrics
-# | fig-cap: "Clustering Evaluation Metrics by Model and k"
-# | fig-subcap: ["Silhouette Score", "Davies–Bouldin Index", "Calinski–Harabasz Score"]
-# | fig-width: 6.5
-# | fig-height: 5
+# | fig-cap: Clustering Evaluation Metrics ($K$-means vs. Gaussian)
+# | fig-cap-location: top
 # | echo: false
@@
 #### Compute Evaluation Metrics ####
-evals = []
+evaluation = []
 for model_type in ["KMeans", "GMM"]:
@@
         if model_type == "KMeans":
+            # KMeans initialization:
+            # n_clusters: number of clusters
+            # random_state=42: seed for reproducibility
+            # n_init=10: run 10 random initializations and pick the best
             m = KMeans(n_clusters=k, random_state=42, n_init=10)
         else:
+            # GaussianMixture initialization:
+            # n_components=k: number of mixture components (clusters)
+            # random_state=42: seed for reproducibility
             m = GaussianMixture(n_components=k, random_state=42)
         lbls = m.fit_predict(X_scaled)
-        evals.append(
+        evaluation.append(
             {
                 "Model": model_type,
-                "k": k,
+                "$k$": k,
                 "Silhouette": silhouette_score(X_scaled, lbls),
@@
 titles = [
-    "Silhouette Score (↑ better sep.)",
-    "Davies–Bouldin Index (↓ less overlap)",
-    "Calinski–Harabasz Score (↑ tighter clusters)",
+    "Silhouette Score",
+    "Davies–Bouldin Index",
+    "Calinski–Harabasz Score",
 ]
@@
     for j, model_type in enumerate(["KMeans", "GMM"]):
-        scores = [e[metric] for e in evals if e["Model"] == model_type]
+        scores = [e[metric] for e in evaluation if e["Model"] == model_type]
         ax.bar(x + (j - 0.5) * width, scores, width, label=model_type)
     ax.set_xticks(x)
-    ax.set_xticklabels(["k=2", "k=3"])
+    ax.set_xticklabels(["$k = 2$", "$k = 3$"])
     ax.set_title(titles[idx])
-    ax.set_xlabel("Clusters (k)")
+    ax.set_xlabel("Clusters ($k$)")
     ax.set_ylabel(metric)
@@
 
-fig.suptitle("Clustering Metrics: KMeans vs. GMM", fontsize=16)
 fig.tight_layout(rect=[0, 0.03, 1, 0.95])
@@
 
-@fig-cluster_metrics
+*Note.* @fig-crime_trends shows three cluster‐validation metrics for both $K$-Means and Gaussian Mixture Models at $k = 2$ and $k = 3$.
+For $k = 2$, GMM produces a Silhouette Score of 0.141, Davies–Bouldin Index of 1.958, and Calinski–Harabasz Score of 30.4, whereas $K$-means achieves 0.429, 0.860, and 152.7, respectively.
+Increasing to $k = 3$ improves separation for GMM (Silhouette = 0.345; DB = 0.998; CH = 114.1) but slightly reduces cohesion for $K$-means (Silhouette = 0.424; DB = 0.786; CH = 133.0).
+
+### Average Yearly Crime Rate Change by Neighbourhood
+
+@tbl-rate_change_summary reports the annual percent change in mean crime rates per 100,000 persons for each SES cluster.
+
+```{python}
+# | label: tbl-rate_change_summary
+# | tbl-cap: SES Cluster Crime Rate Change
+# | tbl-cap-location: top
+# | echo: false
+
+#### Yearly crime rate change ####
+# Select and rename columns
+change_summary = rates_change.select(
+    [
+        pl.col("year").alias("Year"),
+        pl.col("crime").alias("Crime"),
+        pl.col("High Opportunity_pct_change").alias("High Opportunity (%)"),
+        pl.col("Medium Opportunity_pct_change").alias("Medium Opportunity (%)"),
+        pl.col("Low Opportunity_pct_change").alias("Low Opportunity (%)"),
+    ]
+)
+
+#### Display ####
+change_summary
+```
+
+*Note.* @tbl-rate_change_summary reports the annual percent change in crime rates for each SES cluster.
+Positive values indicating increases and negative values indicating decreases relative to the prior year.
+**Needs re-shaping**.
 

### 2025-05-28 scripts/01-simulated_data_test.py

@@ -44,7 +44,7 @@ def test_column_count(sim_data):
 # Check for missing values
 # [https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.null_count.html]
 def test_no_missing(sim_data):
-    # sum of null counts across all cols should be zero
+    # sum of null counts across all columns should be zero
     total_nulls = sim_data.null_count().to_series().sum()
     assert total_nulls == 0, f"Found {total_nulls} missing values"
 
@@ -90,14 +90,14 @@ def test_non_negative_counts(sim_data):
 
 # Check that the crime rates are actually plausible (guesstimating 0-2500 per 100K people)
 def test_plausible_rates(sim_data):
-    rate_cols = [
+    rate_columns = [
         "assault_rate",
         "robbery_rate",
         "breakenter_rate",
         "homicide_rate",
         "shooting_rate",
     ]
-    for column in rate_cols:
+    for column in rate_columns:
         # Select the min-max aggregates from the column
         min, max = sim_data.select(
             pl.col(column).min().alias(f"{column}_min"),

### 2025-05-28 scripts/03.0-clean_crime_data.py

@@ -105,4 +105,4 @@ clean_df = clean_df.with_columns(
 clean_df = clean_df.rename({c: c.lower() for c in clean_df.columns})
 
 #### Save data ####
-clean_df.write_csv("data/02-analysis_data/analysis_data_crime.csv")
+clean_df.write_csv("data/02-analysis_data/00-analysis_data_crime.csv")

### 2025-05-28 scripts/03.1-clean_profile_data.py

@@ -113,4 +113,4 @@ profile_clean = profile_clean.with_columns(
 )
 
 #### Save data ####
-profile_clean.write_csv("data/02-analysis_data/analysis_data_profiles.csv")
+profile_clean.write_csv("data/02-analysis_data/01-analysis_data_profiles.csv")

### 2025-05-28 scripts/04-merge_crime_profile.py

@@ -14,8 +14,8 @@ import functools  # inherent to Python
 import operator  # inherent to Python
 
 #### Load and merge neighbourhood crime and profile data ####
-crime_df = pl.read_csv("data/02-analysis_data/analysis_data_crime.csv")
-profile_df = pl.read_csv("data/02-analysis_data/analysis_data_profiles.csv")
+crime_df = pl.read_csv("data/02-analysis_data/00-analysis_data_crime.csv")
+profile_df = pl.read_csv("data/02-analysis_data/01-analysis_data_profiles.csv")
 
 # Anti-join to identify any name mismatches; "which crime names do not appear in profile_df?"
 # [https://docs.pola.rs/user-guide/transformations/joins/#semi-join]
@@ -112,4 +112,4 @@ clean_df = merged_df.select(
 )
 
 #### Save data ####
-clean_df.write_csv("data/02-analysis_data/merged_data.csv")
+clean_df.write_csv("data/02-analysis_data/02-analysis_data_merged.csv")

### 2025-05-28 scripts/05-merged_test.py

@@ -16,10 +16,10 @@ import pytest  # test functions across any .py ending with "test"
 
 
 #### Test data ####
-# Test if data loads correctly (uses fixture function to try to read merged_data.csv)
+# Test if data loads correctly (uses fixture function to try to read 02-analysis_data_merged.csv)
 @pytest.fixture
 def merged_data():
-    return pl.read_csv("data/02-analysis_data/merged_data.csv")
+    return pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
 
 
 # Check that the dataset has 158 rows (there are 158 neighbourhoods in Toronto; height in polars)

### 2025-05-28 scripts/06-eda_neighbourhood_clusters.py

@@ -25,13 +25,11 @@ from sklearn.metrics import (
     silhouette_score,
 )  # Measure optimal k via silhouette score (higher is better)
 
-
 # Set random seed for reproducibility
 np.random.seed(838)
 
-
 #### Load SES features ####
-profiles = pl.read_csv("data/02-analysis_data/merged_data.csv")
+profiles = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
 ses_columns = [
     "education_rate",  # proportion adults with a bachelor’s degree or higher
     "prop_single_parent",  # proportion of single-parent households
@@ -122,8 +120,8 @@ print(clustered)
 
 #### Save cluster data ####
 profiles.select(["neighbourhood", "cluster", "Socioeconomic Index"]).write_csv(
-    "data/02-analysis_data/neighbourhood_clusters.csv"
+    "data/02-analysis_data/03-cluster_neighbourhoods.csv"
 )
 
 # Append cluster info back to merged_data
-profiles.write_csv("data/02-analysis_data/merged_data.csv")
+profiles.write_csv("data/02-analysis_data/02-analysis_data_merged.csv")

### 2025-05-28 scripts/07-table_crime_clusters.py

@@ -1,46 +1,83 @@
 #### Preamble ####
-# Purpose: Tables 2019–2024 crime trend data by Toronto neighbourhood clusters.
+# Purpose: Tables 2019–2024 crime trend data by Toronto neighbourhood clusters (percent change).
 # Author: Mike Cowan
 # Date: 19 May 2025
 # Contact: m.cowan@mail.utoronto.ca
 # License: MIT
 # Pre-requisites:
-# - `polars` must be installed (pip install polars)
+# - `polars` must be installed (`pip install polars`)
 
 #### Workspace setup ####
 import polars as pl
-import itertools  # inherent to Python
+import itertools  # for generating crime × year combinations
 
 #### Load data ####
-merged_data = pl.read_csv("data/02-analysis_data/merged_data.csv")
+merged_data = pl.read_csv(
+    "data/02-analysis_data/02-analysis_data_merged.csv"
+)  # contains cluster labels and crime-rate columns
 
 # Set parameters
-group_col = "Socioeconomic_index"  # use consistent lowercase naming
+group_columns = "Socioeconomic_index"  # column name for SES cluster grouping
 crime_types = ["assault", "robbery", "breakenter", "shooting"]
-years = range(2019, 2025)
+years = list(range(2019, 2025))  # inclusive range of years
 
-# Collect individual summaries
+# Collect yearly summaries for each crime type and cluster ####
 records = []
 for crime, year in itertools.product(crime_types, years):
-    colname = f"{crime}_rate_{year}"
-    if colname not in merged_data.columns:
-        continue
+    col_name = f"{crime}_rate_{year}"
+    if col_name not in merged_data.columns:
+        continue  # skip if rate column is missing
 
+    # Compute mean rate per 100K residents for each SES cluster
     summary = (
-        merged_data.select([group_col, colname])
-        .group_by(group_col)
-        .agg(pl.mean(colname).alias("avg_rate"))
+        merged_data.select([group_columns, col_name])
+        .group_by(group_columns)
+        .agg(pl.mean(col_name).alias("avg_rate"))  # average rate for the specified year
         .with_columns([pl.lit(crime).alias("crime"), pl.lit(year).alias("year")])
     )
     records.append(summary)
 
-# Combine all
-result_df = pl.concat(records)
 
-# Pivot to wide format
+# Concatenate all summaries into one DataFrame
+def build_result():
+    return pl.concat(records) if records else pl.DataFrame()
+
+
[…]

### 2025-05-28 scripts/08-plot_crime_clusters.py

@@ -14,7 +14,7 @@ import matplotlib.pyplot as plt
 from pathlib import Path  # inherent to Python
 
 #### Load data ####
-merged_data = pl.read_csv("data/02-analysis_data/merged_data.csv")
+merged_data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
 crime_types = ["assault", "breakenter", "robbery", "shooting"]
 years = [2019, 2020, 2021, 2022, 2023, 2024]
 

### 2025-05-28 scripts/09-model_evaluation.py

@@ -27,7 +27,7 @@ from sklearn.metrics import (
 )
 
 # Load and scale SES features (equal weighting requirement for clustering)
-data = pl.read_csv("data/02-analysis_data/merged_data.csv")
+data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
 ses_columns = [
     "education_rate",
     "prop_single_parent",
@@ -41,7 +41,7 @@ scaled_matrix = StandardScaler().fit_transform(feature_matrix)
 # [https://scikit-learn.org/stable/modules/generated/sklearn.decomposition.PCA.html]
 pca_coordinates = PCA(n_components=2).fit_transform(scaled_matrix)
 
-# Define clustering configurations (K-means vs. GMM; k=2 vs. k=3) and colour maps
+# Define clustering configurations (K-means vs. GMM; k=2, 3) and colour maps
 cluster_configs = [
     ("KMeans", 2, "Paired", "KMeans (k=2)"),
     ("GMM", 2, "Accent", "GMM (k=2)"),
@@ -77,7 +77,7 @@ for idx, (model_type, num_clusters, colormap, title) in enumerate(cluster_config
     axes_pca[idx].legend(*scatter.legend_elements(), title="Cluster")
 
 fig_pca.suptitle(
-    "Dimensionality Reduced Clustering Results (KMeans vs. GMM)", fontsize=16
+    "Dimensionality Reduced Clustering Results ($K$-Means vs. GMM)", fontsize=16
 )
 fig_pca.tight_layout(rect=[0, 0.03, 1, 0.95])
 fig_pca.savefig("other/figures/fig_2_cluster_comparisons.png", dpi=300)
@@ -155,7 +155,7 @@ for i, metric in enumerate(metric_names):
 
 #### Save metrics figure ####
 fig_metrics.suptitle(
-    "Clustering Evaluation Metrics (KMeans vs. GMM; k = 2 vs. k = 3)", fontsize=16
+    "Clustering Evaluation Metrics ($K$-Means vs. GMM; k = 2, 3)", fontsize=16
 )
 fig_metrics.tight_layout(rect=[0, 0.03, 1, 0.95])
 fig_metrics.savefig("other/figures/fig_3_cluster_metrics.png", dpi=300)
@@ -177,16 +177,16 @@ eval_table = (
     .rename(
         {
             "Model": "Clustering Algorithm",
-            "k": "Clusters (k)",
[…]

### 2025-05-28 scripts/00-simulate_data.py
Corrected typos.

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of 2021-2024 Toronto neighbourhood crime data merged with Census data on single/dual-family households.
+# Purpose: Simulates a dataset of 2021-2024 Toronto neighbourhood crime data merged with Census data
 # Author: Mike Cowan
 # Date: 19 May 2025
 # Contact: m.cowan@mail.utoronto.ca

### 2025-05-28 scripts/02-download_data.py

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Downloads and saves Toronto neighbourhood profile and crime data from Open Data Toronto CKAN.
+# Purpose: Downloads and saves Toronto neighbourhood profile and crime data from Open Data Toronto's CKAN
 # Author: Mike Cowan
 # Date: 19 May 2025
 # Contact: m.cowan@mail.utoronto.ca

### 2025-05-28 scripts/07-table_crime_clusters.py

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Tables 2019–2024 crime trend data by Toronto neighbourhood clusters (percent change).
+# Purpose: Tables 2019–2024 crimee trend data by Toronto neighbourhood clusters (percent change).
 # Author: Mike Cowan
 # Date: 19 May 2025
 # Contact: m.cowan@mail.utoronto.ca

### 2025-05-28 scripts/09-model_evaluation.py

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Evaluates model fit of Kmeans vs. Gaussian clusters (k = 2, k = 3).
+# Purpose: Evaluates model fit of K-means vs. Gaussian clusters (k = 2, k = 3).
 # Author: Mike Cowan
 # Date: 19 May 2025
 # Contact: m.cowan@mail.utoronto.ca

### 2025-05-29 paper/paper.qmd
Committed final update.

@@
   echo: false
-abstract: This paper examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 159 Toronto neighbourhoods.
+abstract: This paper examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 158 Toronto neighbourhoods.
 Drawing on data from the 2021 Canadian Census and the Toronto Police Service, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (High-, Medium-, and Low- Opportunity) and compared crime rates across these clusters over time.
@@
 import matplotlib.font_manager as fm
+import pandas as pd
 from pathlib import Path
@@
 #### Standardize Figure Font in matplotlib ####
+# CM Serif downloaded to ../paper/fonts/cmunrm.otf
 # [https://matplotlib.org/stable/users/explain/text/usetex.html]
@@
 merged_data = pl.read_csv("../data/02-analysis_data/02-analysis_data_merged.csv")
-rates_change = pl.read_csv("../data/02-analysis_data/04-cluster_crime_rates.csv")
+assault_change = pl.read_csv("../data/03-table_data/assault_rate_change.csv")
+breakenter_change = pl.read_csv("../data/03-table_data/breakenter_rate_change.csv")
+robbery_change = pl.read_csv("../data/03-table_data/robbery_rate_change.csv")
+shooting_change = pl.read_csv("../data/03-table_data/shooting_rate_change.csv")
 
@@
 
-#### PCA Coordinates ####
+#### Map PCA Coordinates ####
 pca_coords = PCA(n_components=2).fit_transform(X_scaled)
@@
 cluster_configs = [
-    ("KMeans", 2, "Paired", "KMeans (k=2)"),
-    ("GMM", 2, "Accent", "GMM (k=2)"),
-    ("KMeans", 3, "Set1", "KMeans (k=3)"),
+    ("KMeans", 2, "Paired", "$K$-means ($k$ = 2)"),
+    ("GMM", 2, "Accent", "GMM ($k$ = 2)"),
+    ("KMeans", 3, "Set1", "$K$-means ($k$ = 3)"),
     ("GMM", 3, "Set2", "GMM (k=3)"),
@@
 This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods diverge sharply in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage [@jargowsky_2015_concentrated].
-A plethora of research confirms that these inequalities stratify crime patterns, making city-wide or national averages poor proxies for local realities [@wang_2019_the; @uesugi_2024_a; @yu_2025_how].
 
-Building on this established research, this paper examines how Toronto’s neighbourhood-level crime patterns evolved during the pandemic and its aftermath.
-We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities—often termed “opportunity” or “advantage” differences [@yu_2025_how]—relate to diverging crime trends.
+To further examine these phenomena, this paper analyzes Toronto’s neighbourhood-level crime trends during the pandemic and its aftermath.
+We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities relate to diverging crime trends.
 Specifically, we classify Toronto neighbourhoods into Low-, Medium-, and High-Opportunity clusters via $K$-means clustering on socioeconomic proxies and compare the crime trajectories of these clusters over time.
 
-This paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends.
-Next, we describe our data and methodological approach, justifying our construction of neighbourhood opportunity clusters and analyzing their distinct crime trajectories from 2019-2024.
-Practically, we consider what our neighbourhood-level results suggest for urban crime policy.
+This paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends (@sec-literature-review).
+Next, we describe our data (@sec-data) and model (@sec-model), justifying our construction of neighbourhood opportunity clusters and analyzing their distinct crime trajectories from 2019 to 2024 (@sec-results, with model diagnostics provided in @sec-model-evaluation).
+We then consider what our neighbourhood-level results might suggest for understanding spatially differentiated crime dynamics in Toronto (@sec-discussion).
+We conclude by discussing study limitations (@sec-limitations).
 
-# Literature Review
+# Literature Review {#sec-literature-review}
 
@@
 Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls.
-The resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship and institutional neglect [@antunes_2021_social; @frevel_2021_local].
-Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies particularly during systemic crises-e.g., the COVID-19 pandemic [@antunes_2021_social].
+The resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship-an institutional presence to deter crime-and institutional neglect [@antunes_2021_social; @frevel_2021_local].
+Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies, particularly during systemic crises-e.g., the COVID-19 pandemic [@antunes_2021_social].
 
-Contemporary research extends these theories by emphasizing the study of spatial inequality—the juxtaposition of affluence and deprivation across proximate neighbourhoods.
-For example, Kang (2016) demonstrates that inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence, suggesting that proximity to wealth exacerbates perceptions of exclusion and strain.
-These findings align with contemporary research in Paterson, NJ, which suggested median household income exerted the strongest neighbourhood-level influence on crime patterns [@yu_2025_how's].
+More recent work highlights the importance of spatial inequality—the juxtaposition of affluence and deprivation across proximate neighbourhoods; inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence (Kang, 2016)-suggesting that proximity to wealth exacerbates perceptions of exclusion and strain [@wang_2019_the; @uesugi_2024_a].
+These findings align with contemporary research in Paterson, NJ, which suggests that median household income exerts the strongest neighbourhood-level influence on crime patterns [@yu_2025_how].
 
-Unsurprisingly, decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@frevel_2021_local; @jargowsky_2015_concentrated].
+Decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@frevel_2021_local; @jargowsky_2015_concentrated].
 Toronto exemplifies this dynamic: neighbourhoods with higher marginalization indices exhibit disproportionately high rates of violent and property crime [@wang_2019_the], while areas with entrenched poverty report elevated homicide incidence [@mohammadi_2022_homicide].
 However, the spatial heterogeneity of these relationships in major cities is also critical.
-For example, geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods, demonstrating the need for crime models that are sensitive to local socioeconomic dynamics.
+For example, geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods.
 
-The COVID-19 pandemic undoubtedly exposed and amplified these spatial inequities.
-Initial lockdowns reduced city-wide crime through diminished routine activities, but declines in some cities were uneven [@andresen_2022_in].
+The COVID-19 pandemic exposed and amplified these spatial inequities.
+Initial lockdowns reduced city-wide crime by diminishing routine activities, but declines in some cities were uneven [@andresen_2022_in].
 For instance, while property crime in Vancouver decreased in wealthy neighbourhoods with robust security infrastructure, violent crime surged in disadvantaged areas strained by disrupted social services and weakened guardianship.
 
-These disparities underscore three critical lessons: first, structural disadvantage predicts vulnerability to crime spikes during crises, aligning with social disorganization theory’s emphasis on resource-deprived communities.
+These patterns underscore three lessons: First, structural disadvantage predicts vulnerability to crime spikes during crises, aligning with social disorganization theory’s emphasis on resource-deprived communities.
 Second, inter-neighbourhood inequality magnifies criminogenic risks, as strain theory posits when relative deprivation fuels frustration.
-Third, opportunity structures—access to security, institutional support, and economic stability—can be spatially stratified, privileging affluent areas with systemic advantages.
+Third, opportunity structures—e.g., access to security, institutional support, and economic stability—can be spatially stratified, privileging affluent areas with systemic advantages.
 
-Building on these insights, we operationalize “opportunity” as relative access to stabilizing socioeconomic conditions—a small composite measure proxied by median income, educational attainment, employment status, and household structure.
+Building on these insights, we operationalize “opportunity” as relative access to stabilizing socioeconomic conditions—a small, composite measure proxied by median income, educational attainment, employment status, and household structure.
 Our analysis, focusing on Toronto neighbourhoods, compliments prior research in two ways: first, we expand the temporal scope to 2019–2024, assessing whether pandemic-era disparities persisted into the recovery phase.
@@
 
-We use the programming language Python [@python] alongside polars [@polars] and sci-kit-learn [@scikit-learn] for the paper's data cleaning and analysis.\footnote{Workflow and GitHub-based project structure follow best practices guided by Alexander (2023), which has proven to be an excellent resource: [https://tellingstorieswithdata.com]}.
+We utilize the programming language Python [@python] alongside Polars [@polars] and scikit-learn [@scikit-learn] for the paper’s data cleaning and analysis.\footnote{The workflow and GitHub-based project structure follow best practices guided by Alexander (2023), which has proven to be an excellent resource: [https://tellingstorieswithdata.com].}
 
@@
 
-We sourced crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, break-and-enters, robberies, and shootings—the types of crime we chose to analyze between 2019 and 2024.
+We sourced crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, break-and-enters (B&E), robberies, and shootings—the types of crime we chose to analyze between 2019 and 2024.
+Crime rates per 100,000 persons were calculated by the City of Toronto using population estimates from Environics Analytics, in accordance with Statistics Canada’s standard definition.
 Each entry includes incident years and total counts by neighbourhood (see @tbl-crime_data_preview).
@@
         # Remove quotes from AREA_NAME values
-        pl.col("_id"),
         pl.col("AREA_NAME").str.slice(0, 16),  # keep only first 16 characters
@@
         pl.col("BREAKENTER_2014").fill_null(0),
-        pl.lit("").alias("Robbery_2014").alias("…"),  # fake truncate
+        pl.lit("").alias("Robbery_2014").alias("…"),  # fake column truncate
     ]
@@
 #### Display ####
-crime_glimpse
+crime_glimpse.to_pandas()
 ```
@@
 ### Neighbourhood Census Profiles
-2021 Toronto Census Profile data was obtained from the Open Data Toronto Portal [@citeOpenDataToronto].
-This data was transformed to align with neighbourhood identifiers used in the crime dataset (see @tbl-census_data_preview).
+
+2021 Toronto Census Profile data was obtained from the Open Data Toronto Portal [@citeOpenDataToronto], which draws on Statistics Canada’s Census of Population to compile selected demographic and socioeconomic indicators.
+We transformed a subset of these variables—originally aggregated to Toronto’s 158 social planning neighbourhoods—to align with the neighbourhood identifiers used in the crime dataset (see @tbl-census_data_preview).
 
@@
             pl.col("Elms-Old Rexdale").alias("Elms-Old"),
-            pl.lit("").alias("…"),  # fake truncate
+            pl.lit("").alias("…"),  # fake column truncate
         ]
@@
 #### Display ####
-profile_glimpse
+profile_glimpse.to_pandas()
 ```
@@
 
-Transforming the lived realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for each neighbourhood and (2) aggregating police‐reported crime counts to rate units.
+Transforming the realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for each neighbourhood and (2) aggregating police‐reported crime counts to rate units.
 
@@
 
-From the crime dataset, we aggregated annual counts of assaults, break-and-enters, robberies and shootings (2019–2024) and transformed them into rates per 100,000 persons using neighbourhood-level population denominators.
+From the crime dataset, we aggregated annual counts of assaults, B&E, robberies, and shootings (2019–2024).
+Where population denominators were available, we converted raw counts into rates per 100,000 persons at the neighbourhood level.
+In instances where rate data were already provided, we retained the available figures without transformation.
 
-These two sources were paired by Toronto’s 159 neighbourhoods, yielding socioeconomic features and crime rates from the early pre- to post-COVID-19 pandemic.
+These two sources were paired by Toronto’s 158 neighbourhoods, yielding socioeconomic features and crime rates from the early pre-pandemic to post-pandemic periods.
 
-# Model
+# Model {#sec-model}
 
-Cognizant of prior criminological studies—which have employed powerful spatial models to characterize neighbourhood heterogeneity in crime and socioeconomic conditions (utilizing substantially more predictors; [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how])—we apply $K$-means clustering to four standardized indicators—education rate, proportion of single-parent households, unemployment rate, and median household income—to cluster neighbourhoods based on their socioeconomic similarity.
+Cognizant of prior criminological studies—which have employed powerful spatial models to characterize neighbourhood heterogeneity in crime and socioeconomic conditions, utilizing substantially more predictors [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how]—we apply $K$-means clustering to four standardized indicators—education rate, proportion of single-parent households, unemployment rate, and median household income—to cluster neighbourhoods based on their socioeconomic similarity.
 
-The $K$-means algorithm identifies the set of cluster centroids ${\mu_k}$ that minimizes the total within-cluster sum of squared distances (WCSS)\footnote{Mathematicals have been adapted from Hastie et al.
+The $K$-means algorithm identifies the set of cluster centroids ${\mu_k}$ that minimizes the total within-cluster sum of squared distances (WCSS)\footnote{Mathematical notation adapted from Hastie et al.
 (2009).}.
@@
 
-Let $\mathbf{x}i = (x{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, where $i = 1, \dots, N = 159$.
+Let $\mathbf{x_i} = (x_{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, where $i = 1, \dots, N = 158$.
 Each vector includes standardized values for (1) median household income, (2) the proportion of residents with a bachelor’s degree or higher, (3) the unemployment rate, and (4) the single-parent household rate.
 
-Each raw feature os standardized by 
+Each raw feature is standardized by 
 $$
@@
 # | label: tbl-cluster_stats
-# | tbl-cap: Descriptive Statistics by SES Cluster (2019–2024)
+# | tbl-cap: Descriptive Statistics by Opportunity Cluster (2019–2024)
 # | tbl-cap-location: top
-# | echo: false
 
@@
         [
-            pl.count("neighbourhood").alias("Neighbourhoods (n)"),
-            pl.mean("median_income").alias("Avg. Median Income ($)"),
-            pl.mean("prop_single_parent").round(2).alias("Single-Parent Proportion"),
-            (pl.mean("education_rate")).round(2).alias("Education Rate"),
-            (pl.mean("unemployment_rate")).round(2).alias("Unemployment Rate"),
+            pl.count("neighbourhood").alias("Neighbourhoods"),
+            pl.mean("median_income").round(2).alias("Med. Income ($)"),
+            pl.mean("prop_single_parent").round(2).alias("Single-Parent"),
+            pl.mean("education_rate").round(2).alias("Education"),
+            pl.mean("unemployment_rate").round(2).alias("Unemployed (%)"),
         ]
@@
 Low-Opportunity (Cluster 2) neighbourhoods are characterized by lower education and income, and higher unemployment and single-parent household rates.
-All values represent means of each indicator across the study period.
-**Needs reshaping**
+All values represent means of each indicator across the period of study.
 
-Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating a mean vector and covariance matrix for each component.}—which relax the spherical cluster assumption imposed by $K$-means [@scikit-learn—we], we retained the $K$-means technique due to its superior evaluation scores and greater interpretability (both methods produced broadly similar neighbourhood groupings; see @fig-cluster_pca).
+Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating a mean vector and covariance matrix for each component.}—which relax the spherical cluster assumption imposed by $K$-means [@scikit-learn], we retained the $K$-means technique due to its superior evaluation scores and greater interpretability (both methods produced broadly similar neighbourhood groupings; see @fig-cluster_pca).
 
-# Results
+# Results {#sec-results}
 
@@
 # | fig-cap-location: top
-# | echo: false
 
@@
 
-# Loop through each crime type and plot trends by SES cluster #
+# Loop through each crime type and plot trends by Opportunity cluster #
 for idx, crime in enumerate(crime_types):
@@
     merged_long = (
-        merged_data.select(
-            ["Socioeconomic_index", *rate_cols])
+        merged_data.select(["opportunity_index", *rate_cols])
         .unpivot(
             on=rate_cols,
-            index="Socioeconomic_index",
+            index="opportunity_index",
             variable_name="year",
@@
     trend = (
-        merged_long.group_by(["Socioeconomic_index", "year"])
+        merged_long.group_by(["opportunity_index", "year"])
         .agg(pl.col("rate").mean().alias("average_rate"))
-        .sort(["Socioeconomic_index", "year"])
+        .sort(["opportunity_index", "year"])
     )
@@
     for group in ["Low Opportunity", "Medium Opportunity", "High Opportunity"]:
-        sub = trend.filter(pl.col("Socioeconomic_index") == group)
+        sub = trend.filter(pl.col("opportunity_index") == group)
         ax.plot(
@@
     ax.set_ylabel("Rate per 100K persons")
-    ax.legend(title="SES Cluster", fontsize=8)
+    ax.legend(title="Opportunity Cluster", fontsize=8)
 
@@
 
-Figure @fig-crime_trends illustrates how crime unfolded across Toronto’s neighbourhoods between 2019 and 2024.
+@fig-crime_trends illustrates how crime unfolded across Toronto’s neighbourhoods between 2019 and 2024.
 In Low-Opportunity neighbourhoods, assault rates edged upward from 758 to 805 incidents per 100,000 persons—an increase of 6.2%—while Medium-Opportunity areas saw an 8.0% rise, from 737 to 796 per 100,000.
@@
 
-By contrast, property-type and violent crimes more pronounced divergence.
-Robbery rates in High-Opportunity areas plunged by roughly one-third (a 34.5% reduction), falling from 95 to 62 per 100,000, whereas Low-Opportunity clusters recorded a smaller 16.9% decline (126 → 104).
+By contrast, the divergence between neighbourhood type and violent crimes is more pronounced.
+Robbery rates in High-Opportunity areas decreased by roughly one-third (a 34.5% reduction), falling from 95 to 62 per 100,000, whereas Low-Opportunity clusters recorded a smaller decline of 16.9% (126 → 104).
 Medium-Opportunity wards registered an 18.5% drop (124 → 101).
@@
 
-The 2020 city-wide dip in break-and-enters—driven, in part, by greater residential presence—was particularly deep in deprived areas, where rates fell 26.0% from 246 to 182 per 100,000 between 2019 and 2020.
-However, by 2023 Low-Opportunity wards had rebounded to 214 incidents per 100,000—just 12.9% below their pre-pandemic level.
-Medium-Opportunity neighbourhoods saw a more sustained decline, with break-and-enters down 24.1% (329 → 249), while High-Opportunity areas only edged 2.2% above 2019 levels (348 → 356)\footnote{See Table 3 for tabular breakdown of year-to-year rate change.}.
+The 2020 city-wide dip in B&Es—driven, in part, by greater residential presence—was particularly deep in deprived areas, where rates fell 26.0% from 246 to 182 per 100,000 between 2019 and 2020.
+However, by 2023, Low-Opportunity wards had rebounded to 214 incidents per 100,000—just 12.9% below their pre-pandemic level.
+Medium-Opportunity neighbourhoods saw a more sustained decline, with B&E down 24.1% (329 → 249), while High-Opportunity areas only edged 2.2% above 2019 levels (348 → 356)\footnote{See Table 3 for the breakdown of the year-to-year rate change.}.
 
-# Discussion
+# Discussion {#sec-discussion}
 
-The patterns we observed across Toronto’s neighbourhoods from 2019 to 2024 align closely with the expectations derived from social disorganization and strain theories.
-As predicted, Low Opportunity areas—characterized by lower income, education, and employment levels, and higher single-parent prevalence—exhibited consistently elevated crime rates across all four crime types.
-Expectedly, these neighbourhoods began the pre-pandemic era with higher baseline crime levels but also demonstrated greater volatility in response to pandemic-related shocks.
+Adopting a three-cluster solution revealed noticeable patterns that would be obscured by a binary split (specifically, $K = 2$; see @sec-model-evaluation), underscoring the value of an intermediate socioeconomic category.
+While Low-Opportunity neighbourhoods began the pre-pandemic period with higher baseline crime as was expected, the Medium-Opportunity cluster exhibited dynamic behaviour, pairing with High- and Low-Opportunity trajectories depending on crime type.
 
-Social disorganization theory helps explain these patterns by highlighting how concentrated disadvantage weakens collective efficacy and informal social control mechanisms [@frevel_2021_local].
-The relatively rapid rebound in property crime and the persistent elevation of violent crime in less advantaged areas suggest that these neighbourhoods may lack the institutional buffers and community cohesion necessary to sustain safety gains during and after the 2020 lockdown.
-In contrast, High Opportunity neighbourhoods—presumably richer in social capital and surveillance infrastructure—remained largely insulated from post-pandemic escalations [@wang_2019_the].
+Low- and Medium-Opportunity areas displayed similar rates of year-to-year change across all crime types (with the exception of B&E): a sharp decline during the 2020 lockdown, followed by a steady rise from 2022 to 2023.
+This suggests a shared vulnerability once a threshold of concentrated disadvantage is crossed, in alignment with social disorganization theory [@antunes_2021_social; @frevel_2021_local].
+Weakened collective efficacy and disrupted informal social controls likely triggered uniform social breakdown across these areas, of their relative positions on the socioeconomic spectrum.
 
-Strain theory offers a complementary lens for interpreting the steeper rise in robbery and shootings in Low Opportunity clusters following 2020.
-These forms of violent and acquisitive crime may reflect cumulative stressors—e.g., economic hardship, institutional disengagement, and perceived exclusion—amplified by the pandemic [@antunes_2021_social].
-The fact that Medium Opportunity areas also experienced upward trends in robbery and shootings suggests a broader diffusion of strain-related pressures, albeit less intensely than has been observed in structurally marginalized zones [@yu_2025_how].
+In contrast, robbery crimes showed a similar arc across all clusters-dipping during the lockdown and rebounding post-2021-but the steepest uptick occurred in Low-Opportunity areas, closely followed by Medium-Opportunity zones.
+This erosion of distinction implies that mounting economic strain and institutional disengagement can rapidly intensify motivations for acquisitive crime in disadvantaged neighbourhoods [@antunes_2021_social].
+Noticeably, High-Opportunity neighborhoods remained relatively insulated, showing only modest growth post-lockdown.
 
-Importantly, the growing gap between Low and Medium clusters—particularly in rates of robbery and gun violence—could indicate a deepening of intra-urban inequality.
-This dynamic resonates with spatial criminological findings that crime concentration is not only persistent but intensifying in the most structurally vulnerable areas [@mohammadi_2022_homicide; @uesugi_2024_a].
+B&E crime rates diverged sharply in the years following the pandemic.
+While all clusters had declined substantially-Medium- and High-Opportunity largely areas moved in tandem-High-Opportunity neighbourhoods experienced a significant spike, surpassing their 2019 rates.
+This spike likely reflects shifts in routine activity patterns: as residents of affluent areas returned to in-person work and travel, homes were left unoccupied more frequently, reducing natural guardianship and increasing opportunities for property crime.
 
-These findings reinforce multiple insights from prior criminology literature: (1) Structural disadvantage is predictive of elevated and volatile crime exposure; (2) Neighbourhoods vary in their capacity to absorb or rebound from systemic shocks [@andresen_2022_in]; (3) Socioeconomic opportunity is not merely correlated with crime rates, but mediates the resilience of communities during periods of crisis [@jargowsky_2015_concentrated, @kang_2016_inequality].
+Lastly, shootings were equally low in Medium-Opportunity and High-Opportunity neighbourhoods, both maintaining relatively flat trajectories from 2019 to 2024.
+In stark contrast, however, Low-Opportunity areas—where the majority of the crimes appear to have occurred—experienced a sharp surge in 2023.
 
-Taken together, the results underscore the importance of continuing to analyze urban crime through the dual lenses of structural inequality and spatial heterogeneity.
-The typological clustering approach employed here—grounded in census-derived socioeconomic indicators—provides a simple, replicable method for the first-pass identification of potential areas of concern, and understanding how patterns of crime map onto entrenched dimensions of social disorganization and strain.
+Social disorganization theory helps to explain the persistent vulnerability of disadvantaged areas [@antunes_2021_social; @frevel_2021_local].
+Concentrated disadvantage erodes collective efficacy and informal controls, hindering the ability to sustain safety gains during and after shocks—such as the lockdown [@jargowsky_2015_concentrated].
+
+These theoretical dynamics are reflected in the observed post-pandemic trends.
+The rapid rebound in property crime and sustained elevation of violent crime in Low-Opportunity areas, alongside upward trends in Medium-Opportunity zones (robbery and shootings, albeit less intensely), also could reflect this diffusion of strain-related pressures [@yu_2025_how].
+Furthermore, strain theory clarifies the steep post-2020 rise in robbery and shootings within Low-Opportunity clusters [@antunes_2021_social].
+Cumulative stressors—economic hardship, institutional disengagement, perceived exclusion—amplified by the pandemic likely intensified motivations for these crimes.
+
+The widening gap between Low- and High-Opportunity clusters, especially for robbery and gun violence, signals deepening intra-urban inequality and intensifying crime concentration in the most vulnerable areas [@mohammadi_2022_homicide; @uesugi_2024_a].
 
@@
 Fourth, our method for grouping neighbourhoods—$K$-means clustering—makes simplifying assumptions.
-It treats all four variables as equally important and assumes the clusters are roughly circular and evenly sized-while in reality, socioeconomic differences are rarely that tidy (@scikit-learn).
-Alternative clustering methods (e.g., GMM) might better reflect the true structure of the data.
+It treats all four variables as equally important and assumes the clusters are roughly circular and evenly sized when, in reality, socioeconomic differences are rarely tidy [@scikit-learn].
+Alternative clustering methods that do not require an *a priori* estimation of $K$ groups (e.g., MAP-DP; see Raykov et al., 2016) may better ascertain the true structure of the data.
 
@@
 ### Cluster Plots
+
 A choice of socioeconomic index can meaningfully shift the resulting clusters.
 One straightforward approach is to standardize each indicator (e.g., income, education, unemployment) to z-scores and sum them [@vyas_2006_constructing].
-Alternatively, factor analysis, which derives weights from latent dimensions (e.g., “Economic Capital” or “Family Stability”) can better reflect the structure among variables but requires more complex modelling and larger samples.
+Alternatively, factor analysis, which derives weights from latent dimensions (e.g., “Economic Capital” or “Family Stability”), can better reflect the structure among variables but requires more complex modelling and larger samples.
 
-Given our inquiry is exploratory, we opted to employ $K$-Means.
+Given that our inquiry is exploratory, we opted to employ $K$-means.
 @fig-cluster_pca reduces our data to the first two principal components for visualization purposes.
-To guard against algorithmic artifacts, we place $K$-Means (hard, spherical boundaries) alongside Gaussian Mixture Models (soft, ellipsoidal groupings) [@scikit-learn].
-Consistency across both methods strengthens confidence that our identified neighbourhood clusters reflect real socioeconomic patterns (and that we made the right selection).
+To guard against algorithmic artifacts, we place K-means—characterized by hard, spherical boundaries—alongside Gaussian Mixture Models, which allow for soft, ellipsoidal groupings (@scikit-learn).
+Consistency across both models strengthens confidence that the identified neighbourhood clusters reflect meaningful socioeconomic divisions—and that our clustering choice was appropriate.
 
@@
 # | label: fig-cluster_pca
-# | fig-cap: Clustering Techniques Comparison ($k = 2, 3$)
+# | fig-cap: Clustering Techniques Comparison ($K = 2, 3$)
 # | fig-cap-location: top
-# | echo: false
 
@@
 *Note.* @fig-cluster_pca displays the two-dimensional projection of neighbourhoods, coloured by assigned cluster.
-At $k = 3$, GMM’s middle component (Cluster 1) overlaps with the Cluster 2, suggesting less clear separation at three clusters.
-Visually, $K$-Means ($k = 3$) appears to be the superior model choice.
+At $K = 3$, GMM’s middle component (Cluster 1) overlaps with the Cluster 2, suggesting less clear separation at three clusters.
+Visually, $K$-means ($K = 3$) appears to be the superior model choice.
 
@@
 
-@fig-cluster_metrics presents a detailed comparison of clustering quality across $K$-means and Gaussian mixture models.
+@fig-cluster_metrics presents a detailed comparison of clustering quality across $K$-means and Gaussian mixture models\footnote{Code and evaluation metrics more detail can be found in scikit-learn documentation (Pedregosa et al., 2011).}.
 We compute the Silhouette Score (ranging from –1 to 1; higher values indicate better cohesion and separation) to assess how closely each point is matched to its own cluster versus the next best alternative.
 The Davies–Bouldin Index measures the average similarity between each cluster and its most similar counterpart—lower values denote more compact, well-separated clusters.
-Finally, the Calinski–Harabasz Score quantifies the ratio of between-cluster dispersion to within-cluster dispersion—higher scores signify clearer, more distinct grouping.
+Finally, the Calinski–Harabasz Score quantifies the ratio of between-cluster dispersion to within-cluster dispersion—higher scores signify clearer, more distinct groupings.
+
+The clustering plots below help illustrate our choice of $K$-means with $K = 3$, which offers clearer group separation and a more stable cluster structure.
+In the $K$-means panels, the groupings are relatively distinct—particularly at $K = 3$, where clusters are well-separated across the principal component space.
+By contrast, the Gaussian Mixture Model results show less defined boundaries; at both values of $K$ (especially $K = 2$), there is greater overlap between adjacent clusters and less visual separation-aligning with GMM’s lower Silhouette and higher Davies–Bouldin Index values (see @fig-cluster_metrics).
 
@@
     ax = axs[idx]
-    x = np.arange(2)  # k = 2, 3
+    x = np.arange(2)  # K = 2, 3
     width = 0.35
@@
     ax.set_xticks(x)
-    ax.set_xticklabels(["$k = 2$", "$k = 3$"])
+    ax.set_xticklabels(["$K = 2$", "$K = 3$"])
     ax.set_title(titles[idx])
-    ax.set_xlabel("Clusters ($k$)")
+    ax.set_xlabel("Clusters ($K$)")
     ax.set_ylabel(metric)
@@
 
-*Note.* @fig-crime_trends shows three cluster‐validation metrics for both $K$-Means and Gaussian Mixture Models at $k = 2$ and $k = 3$.
-For $k = 2$, GMM produces a Silhouette Score of 0.141, Davies–Bouldin Index of 1.958, and Calinski–Harabasz Score of 30.4, whereas $K$-means achieves 0.429, 0.860, and 152.7, respectively.
-Increasing to $k = 3$ improves separation for GMM (Silhouette = 0.345; DB = 0.998; CH = 114.1) but slightly reduces cohesion for $K$-means (Silhouette = 0.424; DB = 0.786; CH = 133.0).
+Note.
+@fig-cluster_metrics displays three cluster-validation metrics for both $K$-means and Gaussian Mixture Models (GMM) at $K = 2$ and $K = 3$.
+For $K = 2$, GMM yields a Silhouette Score of 0.141, a Davies–Bouldin Index (DBI) of 1.958, and a Calinski–Harabasz Score (CHS) of 30.4, whereas $K$-means achieves 0.429, 0.860, and 152.7, respectively.
+Increasing to $K = 3$ improves separation for GMM (Silhouette = 0.345; DBI = 0.998; CHS = 114.1) but slightly reduces cohesion for $K$-means (Silhouette = 0.424; DBI = 0.786; CHS = 133.0).
 
-### Average Yearly Crime Rate Change by Neighbourhood
+While $K = 2$ yields a slightly higher Calinski–Harabasz Score (152.7 vs.
+133.0), its gain in cluster compactness comes at the expense of interpretability: the two-group solution collapses meaningful socioeconomic variation between middle- and high-income neighbourhoods into a single cluster.
+In practical terms, the difference between the average median income of the two upper clusters ($132,368.42 vs.
+$88,775.00) is substantial and justifies treating them as analytically distinct.
 
-@tbl-rate_change_summary reports the annual percent change in mean crime rates per 100,000 persons for each SES cluster.
+### Average Yearly Crime Rate Change by Neighbourhood Opportunity Cluster
+
+#### Neighbourhood Assault
 
 ```{python}
-# | label: tbl-rate_change_summary
-# | tbl-cap: SES Cluster Crime Rate Change
+# | label: tbl-rate_change_assault
+# | tbl-cap: Assault Rate Change by Cluster
 # | tbl-cap-location: top
-# | echo: false
 
-#### Yearly crime rate change ####
-# Select and rename columns
-change_summary = rates_change.select(
-    [
-        pl.col("year").alias("Year"),
-        pl.col("crime").alias("Crime"),
-        pl.col("High Opportunity_pct_change").alias("High Opportunity (%)"),
-        pl.col("Medium Opportunity_pct_change").alias("Medium Opportunity (%)"),
-        pl.col("Low Opportunity_pct_change").alias("Low Opportunity (%)"),
-    ]
+#### Assault % Change ####
+assault_change = assault_change.rename(
+    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
 )
 
-#### Display ####
-change_summary
+#### Display (in pandas to remove backticks, etc.) ####
+assault_change.to_pandas()
 ```
 
-*Note.* @tbl-rate_change_summary reports the annual percent change in crime rates for each SES cluster.
-Positive values indicating increases and negative values indicating decreases relative to the prior year.
-**Needs re-shaping**.
+*Note.* @tbl-rate_change_assault reports the 2019 baseline assault rate and year-over-year percent changes (2020–2024) for each cluster.
+
+#### Break-and-Enter
+
+```{python}
+# | label: tbl-rate_change_breakenter
+# | tbl-cap: Break-and-Enter Rate Change by Cluster
+# | tbl-cap-location: top
+
+#### Break-and-Enter % Change ####
+breakenter_change = breakenter_change.rename(
+    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+)
+
+#### Display (in pandas to remove backticks, etc.) ####
+breakenter_change.to_pandas()
+```
+
+*Note.* @tbl-rate_change_breakenter reports the 2019 baseline break-and-enter rate and annual percent changes.
+
+#### Robbery
+
+```{python}
+# | label: tbl-rate_change_robbery
+# | tbl-cap: Robbery Rate Change by Cluster
+# | tbl-cap-location: top
+
+#### Robbery % Change ####
+robbery_change = robbery_change.rename(
+    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+)
+
+#### Display (in pandas to remove backticks, etc.) ####
+robbery_change.to_pandas()
+```
+
+*Note.* @tbl-rate_change_robbery reports the 2019 baseline robbery rate and annual percent changes.
+
+#### Shooting
+
+```{python}
+# | label: tbl-rate_change_shooting
+# | tbl-cap: Shooting Rate Change by Cluster
+# | tbl-cap-location: top
+
+#### Shooting % Change ####
+shooting_change = shooting_change.rename(
+    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+)
+
+#### Display (in pandas to remove backticks, etc.) ####
+shooting_change.to_pandas()
+```
+
+*Note.* @tbl-rate_change_shooting reports the 2019 baseline shooting rate and subsequent annual percent changes for each cluster.
 

### 2025-05-29 scripts/00-simulate_data.py

@@ -1,91 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates a dataset of 2021-2024 Toronto neighbourhood crime data merged with Census data
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (pip install polars)
-# - `numpy` must be installed (pip install numpy)
-# - `pytest` must be installed (pip install pytest); run with "pytest -q"
-
-#### Workspace setup ####
-import polars as pl
-import numpy as np
-import pytest
-
-# Create NumPy rng generator
-# [https://numpy.org/doc/stable/reference/random/generator.html#numpy.random.Generator]
-rng = np.random.default_rng(
-    838
-)  # random number generator with seed for reproducibility
-
-#### Simulate Census Data ####
-# 158 Toronto neighbourhoods
-neighbourhoods = [
-    f"neighbourhood {i}" for i in range(1, 159)
-]  # generates a list of neighbourhood names (1-158)
-years = [2019, 2020, 2021, 2022, 2023, 2024]  # list of years for the crime data
-
-# Total families per neighbourhood using the min-max from the real dataset
-# e.g., University = lowest, Glenfield-Jane Heights = highest in the real dataset
-# [https://tellingstorieswithdata.com/21-python_essentials.html#python-vs-code-and-uv]
-total_families = rng.integers(
-    low=600, high=4290, size=158
-)  # rng 158 family counts within the range
-
-# Proportion of single-parent families (Beta distribution for proportions, bounded from ~15-55% based on the real dataset)
-prop_single_parent = np.clip(
-    rng.beta(4, 18, size=158),  # α=4, β=18 (mean ~0.18)
-    0.15,  # force lower bound at 15%
-    0.55,  # force upper bound at 55%
-)
-
-# Family counts
-single_parent_families = (prop_single_parent * total_families).round().astype(int)
-two_parent_families = total_families - single_parent_families
-
-# Total population per neighbourhood using the min-max from the real dataset
-total_population = rng.integers(low=6260, high=33300, size=158)
-
-#### Simulate Crime Data ####
-# Crime counts (Poisson draws = counts) guesstimating baselines (more common crimes have higher rates)
-# e.g., crime_count = rng.poisson(* crimes per 1000 people scaled to neighbourhood size)
[…]

### 2025-05-29 scripts/00.0-run_pipeline.py

@@ -1,0 +1,69 @@
+#### Preamble ####
+# Purpose: Executes the full data processing pipeline for Toronto neighbourhood crime and Census profile data.
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - Imports `importlib` to load and run scripts with prefix-numbered filenames.
+# - Each script must define a `main()` function.
+# - Progress messaging is handled inside each script's `main()`.
+# References:
+# - [https://realpython.com/python-main-function/]
+
+#### Workplace setup ####
+import importlib.util
+from pathlib import Path
+import traceback
+
+import importlib.util  # For loading scripts dynamically
+from pathlib import Path  # For handling file paths
+import traceback  # For printing full error tracebacks
+
+#### Pipeline: ordered list of Python filenames (no need for ".py") ####
+pipeline = [
+    "01.0-simulate_data",
+    "02.0-download_data",
+    "03.0-clean_crime_data",
+    "03.1-clean_profile_data",
+    "04.0-merge_crime_profile",
+    "05.0-eda_neighbourhood_clusters",
+    "06.0-table_crime_clusters",
+    "07.0-plot_crime_clusters",
+    "08.0-model_evaluation",
+]
+
+
+#### Import a Python script by its file path and load as a module ####
+# Module spec: where is the module located, what name to give it and how to load it.
+# [https://docs.python.org/3/library/importlib.html#importlib.util.spec_from_file_location]
+def import_module_from_file(script_file: Path):
+    module_name = script_file.stem.replace(".", "_")  # convert "." to "_"
+    spec = importlib.util.spec_from_file_location(
+        module_name, script_file
+    )  # Create a module spec from the file path
+    module = importlib.util.module_from_spec(spec)
+    spec.loader.exec_module(module)
+    return module
+
+
+#### Main Pipeline Execution ####
+def main():
+    for filename in pipeline:
+        script_path = (
+            Path(__file__).parent / f"{filename}.py"
+        )  # Absolute path to script
+        print(f"Running: {filename}.py")
+
+        try:
+            module = import_module_from_file(script_path)  # Load script as module
+            module.main()  # Call its main() function
+        except Exception:
+            print(f"Error occurred while running: {filename}.py")
+            traceback.print_exc()
+
+
+#### Entry Point ####
+if __name__ == "__main__":
+    main()
+    print("Pipeline completed successfully.")

### 2025-05-29 scripts/01.0-simulate_data.py

@@ -1,0 +1,101 @@
+#### Preamble ####
+# Purpose: Simulates a dataset of 2021-2024 Toronto neighbourhood crime data merged with Census data
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `polars` must be installed (pip install polars)
+# - `numpy` must be installed (pip install numpy)
+# - `pytest` must be installed (pip install pytest); run with "pytest -q"
+
+#### Workspace setup ####
+import polars as pl
+import numpy as np
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Simulating Toronto Census and Neighbourhood Crime Data.")
+
+    #### 0.1.0-simulate_data.py ####
+    #### Simulate Census Data ####
+    # Create NumPy rng generator
+    # [https://numpy.org/doc/stable/reference/random/generator.html#numpy.random.Generator]
+    rng = np.random.default_rng(
+        838
+    )  # random number generator with seed for reproducibility
+    # 158 Toronto neighbourhoods
+    neighbourhoods = [
+        f"neighbourhood {i}" for i in range(1, 159)
+    ]  # generates a list of neighbourhood names (1-158)
+    years = [2019, 2020, 2021, 2022, 2023, 2024]  # list of years for the crime data
+
+    # Total families per neighbourhood using the min-max from the real dataset
+    # e.g., University = lowest, Glenfield-Jane Heights = highest in the real dataset
+    # [https://tellingstorieswithdata.com/21-python_essentials.html#python-vs-code-and-uv]
+    total_families = rng.integers(
+        low=600, high=4290, size=158
+    )  # rng 158 family counts within the range
+
+    # Proportion of single-parent families (Beta distribution for proportions, bounded from ~15-55% based on the real dataset)
+    prop_single_parent = np.clip(
+        rng.beta(4, 18, size=158),  # α=4, β=18 (mean ~0.18)
+        0.15,  # force lower bound at 15%
+        0.55,  # force upper bound at 55%
+    )
+
+    # Family counts
+    single_parent_families = (prop_single_parent * total_families).round().astype(int)
+    two_parent_families = total_families - single_parent_families
+
+    # Total population per neighbourhood using the min-max from the real dataset
+    total_population = rng.integers(low=6260, high=33300, size=158)
+
+    #### Simulate Crime Data ####
[…]

### 2025-05-29 scripts/01.1-simulated_data_test.py

@@ -7,12 +7,12 @@
 # Pre-requisites:
 # - `polars` must be installed (pip install polars)
 # - `pytest` must be installed (pip install pytest); run with "pytest -q"
+# References:
+# - [https://docs.pytest.org/en/stable/]
 
 #### Workspace setup ####
 import polars as pl
 import pytest  # test functions across any .py ending with "test"
-
-# [https://docs.pytest.org/en/stable/]
 
 
 #### Test data ####

### 2025-05-29 scripts/02-download_data.py

@@ -1,55 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves Toronto neighbourhood profile and crime data from Open Data Toronto's CKAN
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `requests` must be installed (pip install requests)
-
-#### Workspace setup ####
-import requests
-
-#### Download data ####
-# Toronto Open Data is stored in a CKAN instance. It's APIs are documented here:
-# [https://docs.ckan.org/en/latest/api/]
-
-# To hit our API, you'll be making requests to:
-base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
-
-# Datasets are called "packages". Each package can contain many "resources"
-# To retrieve the metadata for this package and its resources, use the package name in this page's URL:
-url = base_url + "/api/3/action/package_show"
-
-# Crime package
-params_crime = {"id": "neighbourhood-crime-rates"}
-package_crime = requests.get(url, params=params_crime).json()
-
-# Profile package
-params_profile = {"id": "neighbourhood-profiles"}
-package_profile = requests.get(url, params=params_profile).json()
-
-# To get resource data for crime:
-for c_idx, c_resource in enumerate(package_crime["result"]["resources"]):
-    # for datastore_active resources:
-    if c_resource["datastore_active"]:
-        # To get all records in CSV format:
-        c_url = base_url + "/datastore/dump/" + c_resource["id"]
-        c_resource_dump_data = requests.get(c_url).text
-        #### Save string data to CSV file ####
-        # [https://docs.python.org/3/library/functions.html#open]
-        with open(
-            "data/01-raw_data/neighbourhood_crime.csv", "w", encoding="utf-8"
-        ) as f:
-            f.write(c_resource_dump_data)
-
-# To get resource data for profiles:
-for p_idx, p_resource in enumerate(package_profile["result"]["resources"]):
-    # To get all records in XLSX format:
-    if p_resource.get("format", "").lower() == "xlsx":
-        p_url = p_resource["url"]
-        p_resource_dump_data = requests.get(p_url).content
-        #### Save bytes data to XLSX file ####
-        # [https://docs.python.org/3/library/functions.html#open]
-        with open("data/01-raw_data/neighbourhood_profiles.xlsx", "wb") as f:
-            f.write(p_resource_dump_data)

### 2025-05-29 scripts/02.0-download_data.py

@@ -1,0 +1,67 @@
+#### Preamble ####
+# Purpose: Downloads and saves Toronto neighbourhood profile and crime data from Open Data Toronto's CKAN
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `requests` must be installed (pip install requests)
+
+#### Workspace setup ####
+import requests
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Downloading datasets.")
+
+    #### 02.0-download_data.py ####
+    #### Download data ####
+    # Toronto Open Data is stored in a CKAN instance. It's APIs are documented here:
+    # [https://docs.ckan.org/en/latest/api/]
+
+    # To hit our API, you'll be making requests to:
+    base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
+
+    # Datasets are called "packages". Each package can contain many "resources"
+    # To retrieve the metadata for this package and its resources, use the package name in this page's URL:
+    url = base_url + "/api/3/action/package_show"
+
+    # Crime package
+    params_crime = {"id": "neighbourhood-crime-rates"}
+    package_crime = requests.get(url, params=params_crime).json()
+
+    # Profile package
+    params_profile = {"id": "neighbourhood-profiles"}
+    package_profile = requests.get(url, params=params_profile).json()
+
+    # To get resource data for crime:
+    for c_idx, c_resource in enumerate(package_crime["result"]["resources"]):
+        # for datastore_active resources:
+        if c_resource["datastore_active"]:
+            # To get all records in CSV format:
+            c_url = base_url + "/datastore/dump/" + c_resource["id"]
+            c_resource_dump_data = requests.get(c_url).text
+            #### Save string data to CSV file ####
+            # [https://docs.python.org/3/library/functions.html#open]
+            with open(
+                "data/01-raw_data/neighbourhood_crime.csv", "w", encoding="utf-8"
+            ) as f:
+                f.write(c_resource_dump_data)
+
+    # To get resource data for profiles:
+    for p_idx, p_resource in enumerate(package_profile["result"]["resources"]):
+        # To get all records in XLSX format:
+        if p_resource.get("format", "").lower() == "xlsx":
+            p_url = p_resource["url"]
+            p_resource_dump_data = requests.get(p_url).content
+            #### Save bytes data to XLSX file ####
[…]

### 2025-05-29 scripts/03.0-clean_crime_data.py

@@ -10,99 +10,115 @@
 #### Workspace setup ####
 import polars as pl
 
-#### Load and clean neighbourhood crime data ####
-# Neighbourhood crime data
-crime_df = pl.read_csv("data/01-raw_data/neighbourhood_crime.csv")
 
-# Select columns of interest
-# [https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.select.html#polars.DataFrame.select]
-crime_df = crime_df.select(
-    "AREA_NAME",  # neighbourhoods
-    "ASSAULT_2019",
-    "ASSAULT_RATE_2019",
-    "ASSAULT_2020",
-    "ASSAULT_RATE_2020",
-    "ASSAULT_2021",
-    "ASSAULT_RATE_2021",
-    "ASSAULT_2022",
-    "ASSAULT_RATE_2022",
-    "ASSAULT_2023",
-    "ASSAULT_RATE_2023",
-    "ASSAULT_2024",
-    "ASSAULT_RATE_2024",
-    "BREAKENTER_2019",
-    "BREAKENTER_RATE_2019",
-    "BREAKENTER_2020",
-    "BREAKENTER_RATE_2020",
-    "BREAKENTER_2021",
-    "BREAKENTER_RATE_2021",
-    "BREAKENTER_2022",
-    "BREAKENTER_RATE_2022",
-    "BREAKENTER_2023",
-    "BREAKENTER_RATE_2023",
-    "BREAKENTER_2024",
-    "BREAKENTER_RATE_2024",
-    "HOMICIDE_2019",
-    "HOMICIDE_RATE_2019",
-    "HOMICIDE_2020",
-    "HOMICIDE_RATE_2020",
-    "HOMICIDE_2021",
-    "HOMICIDE_RATE_2021",
-    "HOMICIDE_2022",
-    "HOMICIDE_RATE_2022",
-    "HOMICIDE_2023",
-    "HOMICIDE_RATE_2023",
-    "HOMICIDE_2024",
-    "HOMICIDE_RATE_2024",
-    "ROBBERY_2019",
-    "ROBBERY_RATE_2019",
-    "ROBBERY_2020",
-    "ROBBERY_RATE_2020",
-    "ROBBERY_2021",
-    "ROBBERY_RATE_2021",
-    "ROBBERY_2022",
-    "ROBBERY_RATE_2022",
-    "ROBBERY_2023",
[…]

### 2025-05-29 scripts/03.1-clean_profile_data.py

@@ -10,107 +10,119 @@
 #### Workspace setup ####
 import polars as pl
 
-#### Load and clean neighbourhood profile data ####
-# Neighbourhood profile data
-profile_df = pl.read_excel("data/01-raw_data/neighbourhood_profiles.xlsx")
 
-# Unicode normalization, strip whitespace and replace a problem apostrophe (in Neighbourhood Name)
-# [https://sparkbyexamples.com/polars/strip-entire-polars-dataframe]
-profile_standardize = profile_df.with_columns(
-    pl.col("Neighbourhood Name")
-    .str.normalize(form="NFKC")
-    .str.strip_chars(" ")
-    .str.replace("’", "'")
-    .str.replace("‘", "'")
-    .str.replace("’", "'")
-)
+#### MAIN FUNCTION ####
+def main():
+    print("Cleaning neighbourhood profile data.")
 
-# Filter the specific rows of interest
-profile_filter = profile_standardize.filter(
-    pl.col("Neighbourhood Name").is_in(
+    #### 03.1-clean_profile_data.py ####
+    #### Load and clean neighbourhood profile data ####
+    # Neighbourhood profile data
+    profile_df = pl.read_excel("data/01-raw_data/neighbourhood_profiles.xlsx")
+
+    # Unicode normalization, strip whitespace and replace a problem apostrophe (in Neighbourhood Name)
+    # [https://sparkbyexamples.com/polars/strip-entire-polars-dataframe]
+    profile_standardize = profile_df.with_columns(
+        pl.col("Neighbourhood Name")
+        .str.normalize(form="NFKC")
+        .str.strip_chars(" ")
+        .str.replace("’", "'")
+        .str.replace("‘", "'")
+        .str.replace("’", "'")
+    )
+
+    # Filter the specific rows of interest
+    profile_filter = profile_standardize.filter(
+        pl.col("Neighbourhood Name").is_in(
+            [
+                "Total - Persons in private households - 25% sample data",  # row 37
+                "Couple-family households",  # row 235
+                "One-parent-family households",  # row 238
+                "Median total income of household in 2020 ($)",  # row 245
+                "Unemployment rate",  # row 1972
+                "Total - Highest certificate, diploma or degree for the population aged 25 to 64 years in private households - 25% sample data",  # row 2064
+                "Bachelor's degree or higher",  # row 1992 and 2074 (we want the latter)
+            ]
+        )
[…]

### 2025-05-29 scripts/04-merge_crime_profile.py

@@ -1,115 +1,0 @@
-#### Preamble ####
-# Purpose: Merges the cleaned neighbourhood crime and Census profile data.
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (pip install polars)
-# - `polars` must be installed (pip install polars)
-
-#### Workspace setup ####
-import polars as pl
-import functools  # inherent to Python
-import operator  # inherent to Python
-
-#### Load and merge neighbourhood crime and profile data ####
-crime_df = pl.read_csv("data/02-analysis_data/00-analysis_data_crime.csv")
-profile_df = pl.read_csv("data/02-analysis_data/01-analysis_data_profiles.csv")
-
-# Anti-join to identify any name mismatches; "which crime names do not appear in profile_df?"
-# [https://docs.pola.rs/user-guide/transformations/joins/#semi-join]
-mismatches = crime_df.join(profile_df, on="neighbourhood", how="anti")
-print(mismatches.select("neighbourhood").unique().to_series().to_list())
-
-
-# Merge the two DFs on the neighbourhood column
-# [https://dataguymichael.substack.com/p/the-ultimate-polars-cheat-sheet-for]
-merged_df = crime_df.join(
-    profile_df,
-    on="neighbourhood",
-    how="left",  # keep rows from right dataframe (profile_df)
-)
-
-
-# Reorder columns via select (recommended method for polars); one-parent-family households first, then two-parent families, etc.
-# [https://stackoverflow.com/questions/71353113/polars-how-to-reorder-columns-in-a-specific-order]
-clean_df = merged_df.select(
-    [
-        # profile columns
-        "neighbourhood",
-        "total_households",
-        "two_parent_families",
-        "one_parent_families",
-        "prop_single_parent",
-        "median_income",
-        "unemployment_rate",
-        "total_education",
-        "bachelors_or_higher",
-        "education_rate",
-        # crime columns
-        "assault_2019",
-        "assault_rate_2019",
-        "assault_2020",
-        "assault_rate_2020",
-        "assault_2021",
-        "assault_rate_2021",
-        "assault_2022",
-        "assault_rate_2022",
-        "assault_2023",
-        "assault_rate_2023",
-        "assault_2024",
-        "assault_rate_2024",
[…]

### 2025-05-29 scripts/04.0-merge_crime_profile.py

@@ -1,0 +1,124 @@
+#### Preamble ####
+# Purpose: Merges the cleaned neighbourhood crime and Census profile data.
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `polars` must be installed (pip install polars)
+
+#### Workspace setup ####
+import polars as pl
+import functools  # inherent to Python
+import operator  # inherent to Python
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Merging neighbourhood crime and profile data.")
+
+    #### 04.0-merge_crime_profile.py ####
+    #### Load and merge neighbourhood crime and profile data ####
+    crime_df = pl.read_csv("data/02-analysis_data/00-analysis_data_crime.csv")
+    profile_df = pl.read_csv("data/02-analysis_data/01-analysis_data_profiles.csv")
+
+    # Anti-join to identify any name mismatches; "which crime names do not appear in profile_df?"
+    # [https://docs.pola.rs/user-guide/transformations/joins/#semi-join]
+    mismatches = crime_df.join(profile_df, on="neighbourhood", how="anti")
+    print(mismatches.select("neighbourhood").unique().to_series().to_list())
+
+    # Merge the two DFs on the neighbourhood column
+    # [https://dataguymichael.substack.com/p/the-ultimate-polars-cheat-sheet-for]
+    merged_df = crime_df.join(
+        profile_df,
+        on="neighbourhood",
+        how="left",  # keep rows from right dataframe (profile_df)
+    )
+
+    # Reorder columns via select (recommended method for polars); one-parent-family households first, then two-parent families, etc.
+    # [https://stackoverflow.com/questions/71353113/polars-how-to-reorder-columns-in-a-specific-order]
+    clean_df = merged_df.select(
+        [
+            # profile columns
+            "neighbourhood",
+            "total_households",
+            "two_parent_families",
+            "one_parent_families",
+            "prop_single_parent",
+            "median_income",
+            "unemployment_rate",
+            "total_education",
+            "bachelors_or_higher",
+            "education_rate",
+            # crime columns
+            "assault_2019",
+            "assault_rate_2019",
+            "assault_2020",
+            "assault_rate_2020",
+            "assault_2021",
+            "assault_rate_2021",
+            "assault_2022",
+            "assault_rate_2022",
+            "assault_2023",
+            "assault_rate_2023",
[…]

### 2025-05-29 scripts/04.1-merged_test.py

@@ -7,12 +7,12 @@
 # Pre-requisites:
 # - `polars` must be installed (pip install polars)
 # - `pytest` must be installed (pip install pytest); run with "pytest -q"
+# References:
+# - [https://docs.pytest.org/en/stable]
 
 #### Workspace setup ####
 import polars as pl
 import pytest  # test functions across any .py ending with "test"
-
-# [https://docs.pytest.org/en/stable/]
 
 
 #### Test data ####
@@ -63,7 +63,7 @@ def test_variable_types(merged_data):
         "prop_single_parent": pl.Float64,  # proportion (float between 0 and 1)
         "prop_single_parent": pl.Float64,  # proportion (0–1)
         "median_income": pl.Float64,  # continuous
-        "unemployment_rate": pl.Float64,  # %
+        "unemployment_rate": pl.Float64,  # proportion
         "total_education": pl.Float64,  # count
         "bachelors_or_higher": pl.Float64,  # count
         "education_rate": pl.Float64,  # proportion (0–1)

### 2025-05-29 scripts/05.0-eda_neighbourhood_clusters.py

@@ -1,0 +1,138 @@
+#### Preamble ####
+# Purpose: Generates neighbourhood clusters based on Toronto Census profile data.
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `polars` must be installed (pip install polars)
+# - `numpy` must be installed (pip install numpy)
+# - `matplotlib` must be installed (pip install matplotlib)
+# - `scikit-learn` must be installed (pip install scikit-learn)
+# References:
+# - [https://scikit-learn.org/stable/modules/clustering.html#clustering-evaluation]
+
+#### Workspace setup ####
+import polars as pl
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import (
+    StandardScaler,
+)  # Brings each variable to mean 0 / std 1 so no one feature dominates
+from sklearn.cluster import (
+    KMeans,
+)  # Separate into k groups by minimizing within‐cluster variance
+from sklearn.metrics import (
+    silhouette_score,
+)  # Measure optimal k via silhouette score (higher is better)
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Generating neighbourhood K-means clusters based on Census profile data.")
+
+    #### 05.0-eda_neighbourhood_clusters.py####
+    #### Set random seed for reproducibility ####
+    np.random.seed(838)
+
+    #### Load SES features ####
+    profiles = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
+    ses_columns = [
+        "education_rate",  # proportion adults with a bachelor’s degree or higher
+        "prop_single_parent",  # proportion of single-parent households
+        "unemployment_rate",  # proportioon of the labour force unemployed
+        "median_income",  # median household income
+    ]
+
+    # Fill any missing rate columns with 0.0 (float)
+    crime_types = ["assault", "breakenter", "robbery", "shooting"]
+    years = [2019, 2020, 2021, 2022, 2023, 2024]
+
+    # Create a list of feature columns for the SES and crime rates
+    crime_rate_columns = [
+        f"{crime}_rate_{year}" for crime in crime_types for year in years
+    ]
+
+    # Combine the feature columns with the rate columns
+    profiles = profiles.with_columns(
+        [pl.col(col).fill_null(0.0).alias(col) for col in crime_rate_columns]
[…]

### 2025-05-29 scripts/06-eda_neighbourhood_clusters.py

@@ -1,127 +1,0 @@
-#### Preamble ####
-# Purpose: Generates neighbourhood clusters based on Toronto Census profile data.
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (pip install polars)
-# - `numpy` must be installed (pip install numpy)
-# - `matplotlib` must be installed (pip install matplotlib)
-# - `scikit-learn` must be installed (pip install scikit-learn)
-# Reference: https://scikit-learn.org/stable/modules/clustering.html#clustering-evaluation
-
-#### Workspace setup ####
-import polars as pl
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import (
-    StandardScaler,
-)  # Brings each variable to mean 0 / std 1 so no one feature dominates
-from sklearn.cluster import (
-    KMeans,
-)  # Separate into k groups by minimizing within‐cluster variance
-from sklearn.metrics import (
-    silhouette_score,
-)  # Measure optimal k via silhouette score (higher is better)
-
-# Set random seed for reproducibility
-np.random.seed(838)
-
-#### Load SES features ####
-profiles = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
-ses_columns = [
-    "education_rate",  # proportion adults with a bachelor’s degree or higher
-    "prop_single_parent",  # proportion of single-parent households
-    "unemployment_rate",  # % of the labour force unemployed
-    "median_income",  # median household income
-]
-
-
-# Fill any missing rate columns with 0.0 (float)
-crime_types = ["assault", "breakenter", "robbery", "shooting"]
-years = [2019, 2020, 2021, 2022, 2023, 2024]
-
-# Create a list of feature columns for the SES and crime rates
-crime_rate_columns = [f"{crime}_rate_{year}" for crime in crime_types for year in years]
-
-# Combine the feature columns with the rate columns
-profiles = profiles.with_columns(
-    [pl.col(col).fill_null(0.0).alias(col) for col in crime_rate_columns]
-)
-
-# Convert the selected SES columns to a float array
-X = profiles.select(ses_columns).to_numpy().astype(float)
-
-
[…]

### 2025-05-29 scripts/06.0-table_crime_clusters.py

@@ -1,0 +1,119 @@
+#### Workspace setup ####
+import polars as pl
+import itertools  # for crime-year pairs
+from pathlib import Path
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Generating crime trends by neighbourhood clusters (2019–2024).")
+
+    #### 06.0-table_crime_clusters.py ####
+    #### Load data ####
+    merged_data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
+
+    # Set parameters
+    cluster_col = (
+        "opportunity_index"  # SES cluster label (0 = High, 1 = Medium, 2 = Low)
+    )
+    crime_types = ["assault", "robbery", "breakenter", "shooting"]  # crime categories
+    years = list(range(2019, 2025))  # inclusive year range
+
+    # Collect yearly summaries for each crime type and cluster
+    summaries = []
+    for crime, year in itertools.product(crime_types, years):
+        col_name = f"{crime}_rate_{year}"
+        if col_name not in merged_data.columns:
+            continue  # skip if rate column is missing
+
+        avg_rate = (
+            merged_data.select([cluster_col, col_name])
+            .group_by(cluster_col)
+            .agg(pl.mean(col_name).alias("avg_rate"))
+            .with_columns([pl.lit(crime).alias("crime"), pl.lit(year).alias("year")])
+        )
+        summaries.append(avg_rate)
+
+    # Combine into long format DF
+    all_rates = pl.concat(summaries)
+
+    #### Pivot wider: one row per (crime, year), columns = clusters ####
+    wide_df = all_rates.pivot(
+        index=["crime", "year"], on=cluster_col, values="avg_rate"
+    ).sort(["crime", "year"])
+
+    #### Compute year-to-year percent change for each consecutive year ####
+    cluster_cols = [c for c in wide_df.columns if c not in ["crime", "year"]]
+    pct_change_exprs = []
+    for col in cluster_cols:
+        for prev_year, curr_year in zip(years[:-1], years[1:]):
+            pct_name = f"{col}_pct_{prev_year}_{curr_year}"
+            expr = (
+                pl.when(pl.col("year") == curr_year)
+                .then(
+                    (
+                        (pl.col(col) - pl.col(col).shift(1).over("crime"))
+                        / pl.col(col).shift(1).over("crime")
+                        * 100
+                    ).round(1)
+                )
+                .alias(pct_name)
+            )
[…]

### 2025-05-29 scripts/07-table_crime_clusters.py

@@ -1,83 +1,0 @@
-#### Preamble ####
-# Purpose: Tables 2019–2024 crimee trend data by Toronto neighbourhood clusters (percent change).
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (`pip install polars`)
-
-#### Workspace setup ####
-import polars as pl
-import itertools  # for generating crime × year combinations
-
-#### Load data ####
-merged_data = pl.read_csv(
-    "data/02-analysis_data/02-analysis_data_merged.csv"
-)  # contains cluster labels and crime-rate columns
-
-# Set parameters
-group_columns = "Socioeconomic_index"  # column name for SES cluster grouping
-crime_types = ["assault", "robbery", "breakenter", "shooting"]
-years = list(range(2019, 2025))  # inclusive range of years
-
-# Collect yearly summaries for each crime type and cluster ####
-records = []
-for crime, year in itertools.product(crime_types, years):
-    col_name = f"{crime}_rate_{year}"
-    if col_name not in merged_data.columns:
-        continue  # skip if rate column is missing
-
-    # Compute mean rate per 100K residents for each SES cluster
-    summary = (
-        merged_data.select([group_columns, col_name])
-        .group_by(group_columns)
-        .agg(pl.mean(col_name).alias("avg_rate"))  # average rate for the specified year
-        .with_columns([pl.lit(crime).alias("crime"), pl.lit(year).alias("year")])
-    )
-    records.append(summary)
-
-
-# Concatenate all summaries into one DataFrame
-def build_result():
-    return pl.concat(records) if records else pl.DataFrame()
-
-
-result_df = build_result()
-
-# Pivot_wider: one row per (crime, year) × columns for each SES cluster
-wide_df = result_df.pivot(
-    values="avg_rate", index=["crime", "year"], columns=group_columns
-).sort(["crime", "year"])
-
-# Identify the SES cluster columns for later calculations
-cluster_columns = [c for c in wide_df.columns if c not in ["crime", "year"]]
-
-#### Compute year-to-year percent change for each consecutive year ####
-pct_change_exprs = []
-for col in cluster_columns:
-    # For each target year (excluding the first), compute percent change from the previous year
-    for prev_year, curr_year in zip(years[:-1], years[1:]):
[…]

### 2025-05-29 scripts/07.0-plot_crime_clusters.py

@@ -1,0 +1,111 @@
+#### Preamble ####
+# Purpose: Plots 2019-2024 crime trend data by Toronto neighbourhood clusters.
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `polars` must be installed (pip install polars)
+# - `matplotlib` must be installed (pip install matplotlib)
+
+#### Workspace setup ####
+import polars as pl
+import matplotlib.pyplot as plt
+from pathlib import Path  # inherent to Python
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Plotting crime trends by neighbourhood clusters (2019–2024).")
+
+    #### 07.0-plot_crime_clusters.py ####
+    #### Load data ####
+    merged_data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
+    crime_types = ["assault", "breakenter", "robbery", "shooting"]
+    years = [2019, 2020, 2021, 2022, 2023, 2024]
+
+    # Specify figures directory
+    png_directory = Path("other/figures")
+    png_directory.mkdir(parents=True, exist_ok=True)
+
+    # 2x2 subplot figure for combined trends plot
+    # [https://matplotlib.org/stable/gallery/subplots_axes_and_figures/subplot_demo.html]
+    fig_combined, axes = plt.subplots(2, 2, figsize=(12, 8))
+    axes = axes.flatten()
+
+    # Loop through each crime and save plots
+    for idx, crime in enumerate(crime_types):
+        # Build rate column names for each year
+        rate_columns = [f"{crime}_rate_{y}" for y in years]
+
+        # Unpivot wide to long for time-series, extract year as integer (like pivot_longer)
+        # Before: Columns = years, Rows = neighborhoods
+        # After: Rows = neighborhood-year combinations, Columns = [SES, year, rate]
+        # [https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.unpivot.html#polars.DataFrame.unpivot]
+        merged_long = (
+            merged_data.select(
+                ["opportunity_index", *rate_columns]
+            )  # * = "and all individual columns from list"
+            .unpivot(
+                on=rate_columns,
+                index="opportunity_index",
+                variable_name="year",
+                value_name="rate",
+            )
[…]

### 2025-05-29 scripts/08-plot_crime_clusters.py

@@ -1,98 +1,0 @@
-#### Preamble ####
-# Purpose: Plots 2019-2024 crime trend data by Toronto neighbourhood clusters.
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (pip install polars)
-# - `matplotlib` must be installed (pip install matplotlib)
-
-#### Workspace setup ####
-import polars as pl
-import matplotlib.pyplot as plt
-from pathlib import Path  # inherent to Python
-
-#### Load data ####
-merged_data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
-crime_types = ["assault", "breakenter", "robbery", "shooting"]
-years = [2019, 2020, 2021, 2022, 2023, 2024]
-
-# Specify figures directory
-png_directory = Path("other/figures")
-png_directory.mkdir(parents=True, exist_ok=True)
-
-# 2x2 subplot figure for combined trends plot
-# [https://matplotlib.org/stable/gallery/subplots_axes_and_figures/subplot_demo.html]
-fig_combined, axes = plt.subplots(2, 2, figsize=(12, 8))
-axes = axes.flatten()
-
-
-# Loop through each crime and save plots
-for idx, crime in enumerate(crime_types):
-    # Build rate column names for each year
-    rate_columns = [f"{crime}_rate_{y}" for y in years]
-
-    # Unpivot wide to long for time-series, extract year as integer (like pivot_longer)
-    # Before: Columns = years, Rows = neighborhoods
-    # After: Rows = neighborhood-year combinations, Columns = [SES, year, rate]
-    # [https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.unpivot.html#polars.DataFrame.unpivot]
-    merged_long = (
-        merged_data.select(
-            ["Socioeconomic_index", *rate_columns]
-        )  # * = "and all individual columns from list"
-        .unpivot(
-            on=rate_columns,
-            index="Socioeconomic_index",
-            variable_name="year",
-            value_name="rate",
-        )
-        .with_columns(pl.col("year").str.extract(r"_(\d{4})$", 1).cast(pl.Int64))
-    )
-
-    # Group by SES and year, calculating mean rates
-    trend = (
[…]

### 2025-05-29 scripts/08.0-model_evaluation.py

@@ -1,0 +1,206 @@
+#### Preamble ####
+# Purpose: Evaluates model fit of K-means vs. Gaussian clusters (K = 2, K = 3).
+# Author: Mike Cowan
+# Date: 19 May 2025
+# Contact: m.cowan@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+# - `polars` must be installed (pip install polars)
+# - `numpy` must be installed (pip install numpy)
+# - `matplotlib` must be installed (pip install matplotlib)
+# - `scikit-learn` must be installed (pip install scikit-learn)
+# References:
+# - [https://scikit-learn.org/stable/auto_examples/cluster/plot_kmeans_assumptions.html]
+
+#### Workspace setup ####
+import polars as pl
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.decomposition import PCA
+from sklearn.cluster import KMeans
+from sklearn.mixture import GaussianMixture
+from sklearn.metrics import (
+    silhouette_score,
+    davies_bouldin_score,
+    calinski_harabasz_score,
+)
+
+
+#### MAIN FUNCTION ####
+def main():
+    print("Evaluating K-means vs. Gaussian Mixture clustering models ($K$ = 2, 3).")
+
+    #### 08.0-model_evaluation.py ####
+    # Load and scale SES features (equal weighting requirement for clustering)
+    data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
+    ses_columns = [
+        "education_rate",
+        "prop_single_parent",
+        "unemployment_rate",
+        "median_income",
+    ]
+    feature_matrix = data.select(ses_columns).to_numpy()
+    scaled_matrix = StandardScaler().fit_transform(feature_matrix)
+
+    # PCA for 2D visualization (preserves variance)
+    # [https://scikit-learn.org/stable/modules/generated/sklearn.decomposition.PCA.html]
+    pca_coordinates = PCA(n_components=2).fit_transform(scaled_matrix)
+
+    # Define clustering configurations (K-means vs. GMM; K =2, 3) and colour maps
+    cluster_configs = [
+        ("$K$-means", 2, "Paired", "$K$-means ($K$ = 2)"),
+        ("GMM", 2, "Accent", "GMM ($K$ = 2)"),
[…]

### 2025-05-29 scripts/09-model_evaluation.py

@@ -1,192 +1,0 @@
-#### Preamble ####
-# Purpose: Evaluates model fit of K-means vs. Gaussian clusters (k = 2, k = 3).
-# Author: Mike Cowan
-# Date: 19 May 2025
-# Contact: m.cowan@mail.utoronto.ca
-# License: MIT
-# Pre-requisites:
-# - `polars` must be installed (pip install polars)
-# - `numpy` must be installed (pip install numpy)
-# - `matplotlib` must be installed (pip install matplotlib)
-# - `scikit-learn` must be installed (pip install scikit-learn)
-# References:
-# https://scikit-learn.org/stable/auto_examples/cluster/plot_kmeans_assumptions.html
-
-#### Workspace setup ####
-import polars as pl
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import StandardScaler
-from sklearn.decomposition import PCA
-from sklearn.cluster import KMeans
-from sklearn.mixture import GaussianMixture
-from sklearn.metrics import (
-    silhouette_score,
-    davies_bouldin_score,
-    calinski_harabasz_score,
-)
-
-# Load and scale SES features (equal weighting requirement for clustering)
-data = pl.read_csv("data/02-analysis_data/02-analysis_data_merged.csv")
-ses_columns = [
-    "education_rate",
-    "prop_single_parent",
-    "unemployment_rate",
-    "median_income",
-]
-feature_matrix = data.select(ses_columns).to_numpy()
-scaled_matrix = StandardScaler().fit_transform(feature_matrix)
-
-# PCA for 2D visualization (preserves variance)
-# [https://scikit-learn.org/stable/modules/generated/sklearn.decomposition.PCA.html]
-pca_coordinates = PCA(n_components=2).fit_transform(scaled_matrix)
-
-# Define clustering configurations (K-means vs. GMM; k=2, 3) and colour maps
-cluster_configs = [
-    ("KMeans", 2, "Paired", "KMeans (k=2)"),
-    ("GMM", 2, "Accent", "GMM (k=2)"),
-    ("KMeans", 3, "Set1", "KMeans (k=3)"),
-    ("GMM", 3, "Set2", "GMM (k=3)"),
-]
-
-# Plot PCA visualizations in a 2x2 grid
-fig_pca, axes_pca = plt.subplots(2, 2, figsize=(12, 10))
-axes_pca = axes_pca.flatten()
-
-for idx, (model_type, num_clusters, colormap, title) in enumerate(cluster_configs):
[…]

### 2025-05-31 paper/paper.qmd
Committed update based on feedback.

@@
 ---
-title: "Neighbourhood Crime in Toronto (2019–2024)"
-subtitle: "Analyzing Crime Trends in Low, Medium, and High Socioeconomic Areas"
+title: "The Evolution of Neighbourhood Crime in Toronto (2019–2024)"
+subtitle: "A Cluster Analysis Across Socioeconomic Opportunity Levels"
 author: 
@@
 thanks: |
-  Code and data are available at: [https://github.com/mcowan38/tswd_toronto_crime].
-date: "19 May 2025"
+  Code and data are available at: https://github.com/mcowan38/tswd_toronto_crime.
+date: "today"
 date-format: long
@@
 abstract: This paper examines how neighbourhood-level crime trajectories in Toronto evolved from 2019 through 2024, using socioeconomic clustering to identify crime rate divergence across 158 Toronto neighbourhoods.
-Drawing on data from the 2021 Canadian Census and the Toronto Police Service, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (High-, Medium-, and Low- Opportunity) and compared crime rates across these clusters over time.
-Across all four crime types, Low Opportunity neighbourhoods experienced persistently higher rates—especially for assault and shootings—while High Opportunity areas remained relatively insulated.
-Following a city-wide decline in break-and-enters during the 2020 lockdown, crime rebounded steeply in deprived areas, with robbery and gun violence disproportionately escalating through 2023.
-These findings highlight persistent socioeconomic disparities in crime exposure, with disadvantaged neighbourhoods bearing a disproportionate share of urban crime.
+Drawing on data from the 2021 Canadian Census and the Toronto Police Service, we applied $K$-means clustering to group neighbourhoods into three socioeconomic categories reflecting relative socioeconomic advantage (Low-, Medium-, and High-Opportunity) and compared crime rates across these clusters over time.
+Following the city-wide decline in crime associated with the 2020 lockdown, non-property crime rebounded in the two less-advantaged clusters.
+On aggregate, Low-Opportunity neighbourhoods experienced persistently higher non-property crime rates while High-Opportunity areas remained comparatively insulated from violent crime altogether.
+These findings highlight the socioeconomic disparities in Toronto crime exposure, with disadvantaged neighbourhoods bearing a disproportionate share of urban crime and absorbing most of the post-pandemic increase in crime rates.
 format: 
@@
 The COVID-19 pandemic disrupted social and economic systems worldwide—and crime trends were no exception.
-In Canada, police-reported incidents plummeted in 2020 before rebounding sharply in 2021 (Statistics Canada, 2021), with violent crime surging 5% above pre-pandemic levels while property crime fell to historic lows.
+In Canada, police-reported incidents decreased in 2020 before rebounding in 2021, with violent crime increasing 5% above pre-pandemic levels [@statscan_csi].
 
-However, these national trends obscure local disparities; crime clusters unevenly across urban spaces, concentrated in areas marked by structural inequities—e.g., income inequality, educational disparities, and labour market marginalization [@mohammadi_2022_homicide; @wang_2019_the].
-This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods diverge sharply in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage [@jargowsky_2015_concentrated].
+However, these national trends obscure local disparities; crime is distributed unevenly across urban spaces, concentrating in areas marked by structural inequities—e.g., income inequality, educational disparities, and labour market marginalization [@hipp_2007; @mohammadi_2022; @jargowsky_2015].
+This spatial unevenness is particularly pronounced in Toronto, Canada’s largest city, where neighbourhoods vary widely in socioeconomic conditions—from affluent enclaves to zones of concentrated disadvantage [@wang_2019].
 
-To further examine these phenomena, this paper analyzes Toronto’s neighbourhood-level crime trends during the pandemic and its aftermath.
+To explore these phenomena, this paper traces neighbourhood-level crime trajectories from the pandemic to its aftermath.
 We analyze official crime counts for Toronto’s neighbourhoods from 2019 through 2024, assessing how pre-existing social and economic inequalities relate to diverging crime trends.
@@
 
-This paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends (@sec-literature-review).
-Next, we describe our data (@sec-data) and model (@sec-model), justifying our construction of neighbourhood opportunity clusters and analyzing their distinct crime trajectories from 2019 to 2024 (@sec-results, with model diagnostics provided in @sec-model-evaluation).
+Disaggregating Toronto into three opportunity clusters revealed that crime patterns in Medium-Opportunity neighbourhoods align with either High- or Low-Opportunity neighbourhoods depending on the offence (@sec-results; @sec-cluster-plots).
+For assault and robbery, Medium-Opportunity neighbourhoods tracked Low-Opportunity trajectories: both declined during the 2020 lockdown and rose steadily through 2023, whereas High-Opportunity areas remained comparatively stable.
+Conversely, Medium- and High-Opportunity clusters experienced a tandem decline in break-and-enter (B&E) incidents from 2019 to 2021, although High-Opportunity neighbourhoods experienced a rebound that surpassed pre-pandemic levels in 2023.
+Gun violence stayed consistently low and flat in Medium- and High-Opportunity areas, yet Low-Opportunity neighbourhoods—already recording more than twice as many shootings—registered a renewed increase in 2023 after their lockdown decline.
+
+The paper proceeds as follows: First, we ground our analysis in criminological frameworks—i.e., social disorganization and strain—which help explain how neighbourhood opportunity structures (e.g., income, education, employment) may condition crime trends (@sec-literature-review).
+Next, we describe our data (@sec-data) and model (@sec-model), justifying the construction of neighbourhood opportunity clusters and analyzing their distinct crime trajectories from 2019 to 2024 (@sec-results; model diagnostics provided in @sec-model-evaluation).
 We then consider what our neighbourhood-level results might suggest for understanding spatially differentiated crime dynamics in Toronto (@sec-discussion).
@@
 
-Criminological theory and empirical evidence converge on two insights in urban crime literature: (1) neighbourhood-level structural conditions fundamentally shape crime, and (2) spatial inequalities between neighbourhoods amplify these effects.
+Criminological theory and empirical evidence converge on two principles in the urban crime literature: (1) neighbourhood-level structural conditions fundamentally shape crime, and (2) spatial inequalities between neighbourhoods amplify these effects.
 
-Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls.
-The resulting conditions create criminogenic environments where crime flourishes due to weakened guardianship-an institutional presence to deter crime-and institutional neglect [@antunes_2021_social; @frevel_2021_local].
-Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies, particularly during systemic crises-e.g., the COVID-19 pandemic [@antunes_2021_social].
+Social disorganization theory suggests that poverty, residential instability, ethnic heterogeneity, and family disruption erode collective efficacy—the capacity of communities to enforce informal social controls [@antunes_2021].
+The resulting conditions create criminogenic environments where crime is prevalent due to weakened guardianship—an institutional presence that deters crime—and institutional neglect [@frevel_2021].
+Strain theory complements this perspective, arguing that material deprivation and relative disadvantage generate frustration that may motivate criminal coping strategies during periods of crisis—e.g., the COVID-19 pandemic [@antunes_2021].
 
-More recent work highlights the importance of spatial inequality—the juxtaposition of affluence and deprivation across proximate neighbourhoods; inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence (Kang, 2016)-suggesting that proximity to wealth exacerbates perceptions of exclusion and strain [@wang_2019_the; @uesugi_2024_a].
-These findings align with contemporary research in Paterson, NJ, which suggests that median household income exerts the strongest neighbourhood-level influence on crime patterns [@yu_2025_how].
+Further research highlights the importance of spatial inequality—the juxtaposition of affluence and deprivation across proximate neighbourhoods; inter-neighbourhood income disparities, rather than city-wide inequality alone, predict localized violence [@chamberlain_2015; @kang_2016]—suggesting that proximity to wealth could further exacerbate exclusion and strain [@wang_2019; @uesugi_2024].
+These expectations are consistent with recent evidence from Paterson, New Jersey, where median household income has been identified as the most powerful predictor of crime patterns at the neighbourhood level [@yu_2022].
 
-Decades of evidence confirm that concentrated disadvantage—marked by poverty, unemployment, and single-parent households—correlates strongly with elevated crime rates [@frevel_2021_local; @jargowsky_2015_concentrated].
-Toronto exemplifies this dynamic: neighbourhoods with higher marginalization indices exhibit disproportionately high rates of violent and property crime [@wang_2019_the], while areas with entrenched poverty report elevated homicide incidence [@mohammadi_2022_homicide].
+Decades of evidence confirm that concentrated disadvantage—characterized by poverty, unemployment, and single-parent households—strongly correlates with elevated crime rates [@frevel_2021; @jargowsky_2015; @andresen_2022].
+Toronto exemplifies this dynamic: neighbourhoods with higher marginalization indices exhibit disproportionately high rates of violent and property crime [@wang_2019], while areas with entrenched poverty report elevated homicide incidence [@mohammadi_2022].
 However, the spatial heterogeneity of these relationships in major cities is also critical.
-For example, geographically weighted regression studies in Chicago [@arnio_2012_demography] and Tokyo [@uesugi_2024_a] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods.
+For example, geographically weighted regression studies in Chicago [@arnio_2012] and Tokyo [@uesugi_2024] reveal that the strength—and even the direction—of socioeconomic predictors of burglary or robbery vary across neighbourhoods.
 
 The COVID-19 pandemic exposed and amplified these spatial inequities.
-Initial lockdowns reduced city-wide crime by diminishing routine activities, but declines in some cities were uneven [@andresen_2022_in].
+Initial lockdowns reduced city-wide crime by diminishing routine activities but declines in some cities were uneven [@andresen_2022].
 For instance, while property crime in Vancouver decreased in wealthy neighbourhoods with robust security infrastructure, violent crime surged in disadvantaged areas strained by disrupted social services and weakened guardianship.
 
-These patterns underscore three lessons: First, structural disadvantage predicts vulnerability to crime spikes during crises, aligning with social disorganization theory’s emphasis on resource-deprived communities.
-Second, inter-neighbourhood inequality magnifies criminogenic risks, as strain theory posits when relative deprivation fuels frustration.
-Third, opportunity structures—e.g., access to security, institutional support, and economic stability—can be spatially stratified, privileging affluent areas with systemic advantages.
+To summarize, social disorganization theory suggests that resource deprivation can increase crime risk.
+Inter-neighbourhood inequality—e.g., access to security, institutional support, and economic stability—can be spatially stratified and magnify these risks.
+Strain theory posits that relative deprivation fuels frustration—a motivator for crime.
 
-Building on these insights, we operationalize “opportunity” as relative access to stabilizing socioeconomic conditions—a small, composite measure proxied by median income, educational attainment, employment status, and household structure.
-Our analysis, focusing on Toronto neighbourhoods, compliments prior research in two ways: first, we expand the temporal scope to 2019–2024, assessing whether pandemic-era disparities persisted into the recovery phase.
-Second, we apply $K$-means clustering to classify neighbourhoods into High-, Medium-, and Low-Opportunity clusters using variables (income, education, employment) empirically shown to influence crime patterns [@yu_2025_how].
+Accordingly, we expect that Toronto’s socioeconomically disadvantaged areas were more susceptible to increases in crime rates after the lockdown period.
+To examine this possibility, we operationalize “opportunity” as relative access to stabilizing socioeconomic conditions—a small, composite measure proxied by median income, educational attainment, employment status, and household structure.
+We apply $K$-means clustering to classify neighbourhoods into High-, Medium-, and Low-Opportunity clusters using variables (and proxies) empirically shown to influence crime patterns [i.e., income, education, employment, and household status; @yu_2022].
 If social disorganization and strain theories hold, Low-Opportunity clusters should exhibit both higher baseline crime rates and greater volatility during disruptions, reflecting their structural precarity.
+
+Our analysis complements Toronto municipal research by tracking the persistence of pandemic-era socioeconomic disparities through the recovery phase.
 
@@
 
-We utilize the programming language Python [@python] alongside Polars [@polars] and scikit-learn [@scikit-learn] for the paper’s data cleaning and analysis.\footnote{The workflow and GitHub-based project structure follow best practices guided by Alexander (2023), which has proven to be an excellent resource: [https://tellingstorieswithdata.com].}
+We utilize the programming language Python [@python] alongside Polars [@polars] and scikit-learn [@scikit-learn] for the paper’s data cleaning and analysis.^[The workflow and GitHub-based project structure follow best practices from @tellingstories.]
 
@@
 
-We sourced crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, break-and-enters (B&E), robberies, and shootings—the types of crime we chose to analyze between 2019 and 2024.
+We source crime data from the Open Data Toronto Portal [@citeOpenDataToronto], which includes assaults, B&E, robberies, and shootings—our variables of interest.
 Crime rates per 100,000 persons were calculated by the City of Toronto using population estimates from Environics Analytics, in accordance with Statistics Canada’s standard definition.
@@
 
-2021 Toronto Census Profile data was obtained from the Open Data Toronto Portal [@citeOpenDataToronto], which draws on Statistics Canada’s Census of Population to compile selected demographic and socioeconomic indicators.
+We obtained 2021 Toronto Census Profile data from the Open Data Toronto Portal [@citeOpenDataToronto], which draws on Statistics Canada’s Census of Population to compile selected demographic and socioeconomic indicators.
 We transformed a subset of these variables—originally aggregated to Toronto’s 158 social planning neighbourhoods—to align with the neighbourhood identifiers used in the crime dataset (see @tbl-census_data_preview).
@@
 
-Transforming the realities of Toronto’s neighbourhoods into data suitable for our analysis required two steps: (1) constructing a composite socioeconomic profile for each neighbourhood and (2) aggregating police‐reported crime counts to rate units.
+Transforming Toronto neighbourhood data into an analysis-ready format involved two steps: (1) constructing a composite socioeconomic profile for each neighbourhood (converting counts into rates) and (2) aggregating police‐reported crime rates.
 
-From the Census data, we extracted four attributes that decades of criminological research identify as correlates of neighbourhood crime: median household income, share of adults with a bachelor’s degree or higher, unemployment rate, and proportion of single-parent families.
+From the Census data, we extracted four attributes that years of criminological research identify as correlates of neighbourhood crime: median household income, share of adults with a bachelor’s degree or higher, unemployment rate, and proportion of single-parent families.
+Where population denominators were available, we converted raw counts into neighbourhood proportions.
+In instances where rate data were provided, we retained the available figures without transformation.
 
-From the crime dataset, we aggregated annual counts of assaults, B&E, robberies, and shootings (2019–2024).
-Where population denominators were available, we converted raw counts into rates per 100,000 persons at the neighbourhood level.
-In instances where rate data were already provided, we retained the available figures without transformation.
+From the crime dataset, we aggregated annual rates of assaults, B&E, robberies, and shootings between 2019 and 2024.
 
@@
 
-Cognizant of prior criminological studies—which have employed powerful spatial models to characterize neighbourhood heterogeneity in crime and socioeconomic conditions, utilizing substantially more predictors [@mohammadi_2022_homicide; @uesugi_2024_a; @wang_2019_the; @yu_2025_how]—we apply $K$-means clustering to four standardized indicators—education rate, proportion of single-parent households, unemployment rate, and median household income—to cluster neighbourhoods based on their socioeconomic similarity.
+Cognizant of prior criminological studies—which have employed spatial models to characterize neighbourhood heterogeneity in crime and socioeconomic conditions, utilizing substantially more predictors [@mohammadi_2022; @uesugi_2024; @wang_2019; @yu_2022]—we apply $K$-means clustering to four standardized indicators—education rate, proportion of single-parent households, unemployment rate, and median household income—to cluster neighbourhoods based on their socioeconomic similarity.
 
-The $K$-means algorithm identifies the set of cluster centroids ${\mu_k}$ that minimizes the total within-cluster sum of squared distances (WCSS)\footnote{Mathematical notation adapted from Hastie et al.
-(2009).}.
-In simple terms, the algorithm partitions observations into $K$ groups such that data points within each group are as similar as possible to one another.
+The $K$-means algorithm identifies the set of cluster centroids ${\mu_k}$ that minimizes the total within-cluster sum of squared distances (WCSS)^[Mathematical notation adapted from @hastie_2009.].
+The algorithm partitions observations into $K$ groups such that data points within each group are as similar as possible to one another.
 Similarity is measured by the squared Euclidean distance between each observation and the average of its assigned group—called the cluster centroid.
@@
 
-Let $\mathbf{x_i} = (x_{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of z-scored socioeconomic features for neighbourhood $i$, where $i = 1, \dots, N = 158$.
+Let $\mathbf{x_i} = (x_{i1}, x_{i2}, x_{i3}, x_{i4})$ denote the four-dimensional vector of $z$-scored socioeconomic features for neighbourhood $i$, where $i = 1, \dots, N = 158$.
 Each vector includes standardized values for (1) median household income, (2) the proportion of residents with a bachelor’s degree or higher, (3) the unemployment rate, and (4) the single-parent household rate.
@@
 We tested $K = 2, \dots, 5$ and selected $K = 3$ based on silhouette diagnostics (see @sec-model-evaluation).
-Additionally, Principal Components Analysis—conducted as a diagnostic check—showed that the first two components explain ~89% of total variance, confirming a strong low-dimensional socioeconomic gradient underlying the four indicators.
+Additionally, Principal Components Analysis—conducted as a diagnostic check—showed that the first two components explain ~89% of the total variance, confirming a strong low-dimensional socioeconomic gradient underlying the four indicators.
 
-We adopted a three-group classification—High, Medium, and Low Opportunity—aligned with standard socioeconomic strata commonly used in policy discourse and everyday language (i.e., upper, middle, and lower class).
+We adopted a three-group classification—Low-, Medium-, and Low-Opportunity—aligned with standard socioeconomic strata commonly used in policy discourse and everyday language (i.e., upper, middle, and lower class).
 Based on this scheme, the resulting clusters are labelled as shown in @tbl-cluster_stats below.
+
 ```{python}
@@
 
-*Note.* @tbl-cluster_stats presents the average socioeconomic features by cluster for Toronto neigbourhoods over 2019–2024.
+*Note.* @tbl-cluster_stats presents the average socioeconomic features by cluster for Toronto neighbourhoods over 2019–2024.
 Clusters were derived via $K$-means clustering on standardized education rate, proportion of single-parent households, unemployment rate, and median household income.
@@
 
-Although we also explored Gaussian Mixture Models (GMM)\footnote{GMM models the data as a mixture of $K$ multivariate normal distributions, estimating a mean vector and covariance matrix for each component.}—which relax the spherical cluster assumption imposed by $K$-means [@scikit-learn], we retained the $K$-means technique due to its superior evaluation scores and greater interpretability (both methods produced broadly similar neighbourhood groupings; see @fig-cluster_pca).
+Although we also explored Gaussian Mixture Models (GMM)^[GMM models the data as a mixture of $K$ multivariate normal distributions, estimating a mean vector and covariance matrix for each component.]—which relax the spherical cluster assumption imposed by $K$-means [@scikit-learn]—we retained the $K$-means technique due to its superior evaluation scores and greater interpretability (both methods produced broadly similar neighbourhood groupings; see @fig-cluster_pca).
 
@@
 # | label: fig-crime_trends
-# | fig-cap: Average Crime Rate Trends by Socioeconomic Cluster (2019–2024)
+# | fig-cap: Average Crime Rate Trends by Opportunity Cluster (2019–2024)
 # | fig-cap-location: top
@@
 
-*Note.* @fig-crime_trends plots trends in crime rates for Assault, Break-and-Enter, Robbery, and Shootings.
+*Note.* @fig-crime_trends plots trends in crime rates for Assault, B&E, Robbery, and Shootings.
 For each year and crime type, the figure shows the mean rate across all neighbourhoods within each cluster.
@@
 However, by 2023, Low-Opportunity wards had rebounded to 214 incidents per 100,000—just 12.9% below their pre-pandemic level.
-Medium-Opportunity neighbourhoods saw a more sustained decline, with B&E down 24.1% (329 → 249), while High-Opportunity areas only edged 2.2% above 2019 levels (348 → 356)\footnote{See Table 3 for the breakdown of the year-to-year rate change.}.
+Medium-Opportunity neighbourhoods saw a more sustained decline, with B&E down 24.1% (329 → 249), while High-Opportunity areas only edged 2.2% above 2019 levels (348 → 356)^[See @sec-tables for the breakdown of the year-to-year rate change.].
 
@@
 Adopting a three-cluster solution revealed noticeable patterns that would be obscured by a binary split (specifically, $K = 2$; see @sec-model-evaluation), underscoring the value of an intermediate socioeconomic category.
-While Low-Opportunity neighbourhoods began the pre-pandemic period with higher baseline crime as was expected, the Medium-Opportunity cluster exhibited dynamic behaviour, pairing with High- and Low-Opportunity trajectories depending on crime type.
+While Low-Opportunity neighbourhoods entered the pre-pandemic period with elevated baseline crime rates and, despite some fluctuations, remained the most affected overall, the Medium-Opportunity cluster displayed more variable patterns, aligning with either High- or Low-Opportunity trajectories depending on the crime type.
 
 Low- and Medium-Opportunity areas displayed similar rates of year-to-year change across all crime types (with the exception of B&E): a sharp decline during the 2020 lockdown, followed by a steady rise from 2022 to 2023.
-This suggests a shared vulnerability once a threshold of concentrated disadvantage is crossed, in alignment with social disorganization theory [@antunes_2021_social; @frevel_2021_local].
-Weakened collective efficacy and disrupted informal social controls likely triggered uniform social breakdown across these areas, of their relative positions on the socioeconomic spectrum.
+This suggests a shared vulnerability once a threshold of concentrated disadvantage is crossed, in alignment with social disorganization theory [@antunes_2021; @frevel_2021].
+Weakened collective efficacy and disrupted informal social controls likely triggered uniform social breakdown across these areas of their relative positions on the socioeconomic spectrum.
 
-In contrast, robbery crimes showed a similar arc across all clusters-dipping during the lockdown and rebounding post-2021-but the steepest uptick occurred in Low-Opportunity areas, closely followed by Medium-Opportunity zones.
-This erosion of distinction implies that mounting economic strain and institutional disengagement can rapidly intensify motivations for acquisitive crime in disadvantaged neighbourhoods [@antunes_2021_social].
-Noticeably, High-Opportunity neighborhoods remained relatively insulated, showing only modest growth post-lockdown.
+In contrast, robbery crimes showed a similar arc across all clusters, dipping during the lockdown and rebounding post-2021; however, the steepest uptick occurred in Low-Opportunity neighbourhoods, closely followed by Medium-Opportunity areas.
+This erosion of distinction implies that mounting economic strain and institutional disengagement can rapidly intensify motivations for acquisitive crime in disadvantaged neighbourhoods [@antunes_2021].
+Noticeably, High-Opportunity neighbourhoods remained relatively insulated, showing only modest growth after the lockdown.
 
-B&E crime rates diverged sharply in the years following the pandemic.
-While all clusters had declined substantially-Medium- and High-Opportunity largely areas moved in tandem-High-Opportunity neighbourhoods experienced a significant spike, surpassing their 2019 rates.
-This spike likely reflects shifts in routine activity patterns: as residents of affluent areas returned to in-person work and travel, homes were left unoccupied more frequently, reducing natural guardianship and increasing opportunities for property crime.
+All neighbourhoods experienced a substantial decline in B&E rates in the years following the pandemic.
+However, while Medium- and High-Opportunity areas largely moved in tandem, High-Opportunity neighbourhoods experienced a significant increase in 2023—surpassing their pre-pandemic rates.
+This variation could reflect shifts in routine activity patterns [@wang_2019]: as residents of affluent areas returned to in-person work and travel, homes were left unoccupied more frequently, reducing natural guardianship and increasing opportunities for property crime.
 
 Lastly, shootings were equally low in Medium-Opportunity and High-Opportunity neighbourhoods, both maintaining relatively flat trajectories from 2019 to 2024.
-In stark contrast, however, Low-Opportunity areas—where the majority of the crimes appear to have occurred—experienced a sharp surge in 2023.
+In stark contrast, however, Low-Opportunity areas—where the majority of the crimes appear to have occurred—experienced a substantial increase in 2023.
 
-Social disorganization theory helps to explain the persistent vulnerability of disadvantaged areas [@antunes_2021_social; @frevel_2021_local].
-Concentrated disadvantage erodes collective efficacy and informal controls, hindering the ability to sustain safety gains during and after shocks—such as the lockdown [@jargowsky_2015_concentrated].
+Social disorganization theory offers a useful lens for understanding fluctuations in crime rates post-lockdown.
+Concentrated disadvantage weakens informal social control and collective efficacy, leaving neighbourhoods with fewer resources more susceptible to rising crime when constraints lift [@antunes_2021; @jargowsky_2015].
+Accordingly, Low- and Medium-Opportunity neighbourhoods proved more vulnerable to a post-pandemic rebound in robbery once mobility restrictions eased.
+In contrast, better-resourced neighbourhoods—supported by stronger safety capacities [@frevel_2021]—largely preserved their lockdown-era gains, with the notable exception of burglary.
 
-These theoretical dynamics are reflected in the observed post-pandemic trends.
-The rapid rebound in property crime and sustained elevation of violent crime in Low-Opportunity areas, alongside upward trends in Medium-Opportunity zones (robbery and shootings, albeit less intensely), also could reflect this diffusion of strain-related pressures [@yu_2025_how].
-Furthermore, strain theory clarifies the steep post-2020 rise in robbery and shootings within Low-Opportunity clusters [@antunes_2021_social].
-Cumulative stressors—economic hardship, institutional disengagement, perceived exclusion—amplified by the pandemic likely intensified motivations for these crimes.
+Furthermore, the sustained elevation of violent crime in Low-Opportunity areas, together with a milder rise in Medium-Opportunity zones (for assault and shootings), may indicate the spread of strain-related pressures across disadvantaged neighbourhoods [@antunes_2021].
+Pandemic-amplified stressors—economic hardship, institutional disengagement, and perceived exclusion—likely intensified motivations for these offences.
+These same pressures appear to be widening the crime gap between Low- and High-Opportunity clusters—most clearly in robbery and gun violence—thereby deepening intra-urban inequality and concentrating crime in the city’s most vulnerable areas [@mohammadi_2022; @uesugi_2024; @andresen_2022].
 
-The widening gap between Low- and High-Opportunity clusters, especially for robbery and gun violence, signals deepening intra-urban inequality and intensifying crime concentration in the most vulnerable areas [@mohammadi_2022_homicide; @uesugi_2024_a].
-
-## Limitations {#sec-limitations}
+# Limitations {#sec-limitations}
 
@@
 
-Additionally, the socioeconomic data we use are static, drawn from the 2021 Census.
+Additionally, the socioeconomic variables we use are static, drawn solely from the 2021 Census data.
 These indicators—income, education, unemployment, and single-parent status—cannot capture changes that occurred before or after 2021.
-Moreover, we excluded several potentially important variables—such as neighbourhood ethnic composition, immigration status, and residential mobility—which may shape both crime exposure and neighbourhood opportunity in substantive ways.
+Similarly, we excluded several potentially important variables—such as neighbourhood ethnic composition, immigration status, and residential mobility—which may shape both crime exposure and neighbourhood opportunity in substantive ways [@hipp_2007].
 As a result, our clustering captures only a partial view of structural disadvantage.
@@
 It treats all four variables as equally important and assumes the clusters are roughly circular and evenly sized when, in reality, socioeconomic differences are rarely tidy [@scikit-learn].
-Alternative clustering methods that do not require an *a priori* estimation of $K$ groups (e.g., MAP-DP; see Raykov et al., 2016) may better ascertain the true structure of the data.
+Alternative clustering methods that do not require an *a priori* estimation of $K$ groups [e.g., MAP-DP; @raykov_2016] may better ascertain the true structure of the data.
 
-Future research should address these gaps by using dynamic, multi-year socioeconomic data, incorporating a broader range of variables, and accounting for spatial relationships between neighbourhoods.
+Future research could address these gaps by incorporating a broader range of socioeconomic indicators [see @statscan_cimd], quantifying crime severity rather than relying on incident counts [see @statscan_csi], and accounting for the geospatial relationships between neighbourhoods.
 In addition, combining quantitative patterns with qualitative accounts of neighbourhood residents could also help ground the numbers in the lived experiences of residents.
@@
 
-### Cluster Plots
+The construction of socioeconomic composite measures can meaningfully influence model outcomes [@vyas_2006].
+More sophisticated approaches may more accurately capture the underlying structure among variables, but can require larger samples and introduce greater model complexity.
+Simpler methods, such as standardizing individual indicators (e.g., income, education, unemployment) and summing their $z$-scores offer more straightforward interpretability but may obscure multidimensional relationships.
 
-A choice of socioeconomic index can meaningfully shift the resulting clusters.
-One straightforward approach is to standardize each indicator (e.g., income, education, unemployment) to z-scores and sum them [@vyas_2006_constructing].
-Alternatively, factor analysis, which derives weights from latent dimensions (e.g., “Economic Capital” or “Family Stability”), can better reflect the structure among variables but requires more complex modelling and larger samples.
+Balancing these trade-offs and acknowledging the exploratory scope of our study—alongside a concise, well-theorized set of continuous variables—we chose the $K$-means algorithm for its capacity to generate easily interpretable, mutually exclusive neighbourhood groupings.
 
-Given that our inquiry is exploratory, we opted to employ $K$-means.
+### Cluster Plots {#sec-cluster-plots}
+
 @fig-cluster_pca reduces our data to the first two principal components for visualization purposes.
@@
 
-@fig-cluster_metrics presents a detailed comparison of clustering quality across $K$-means and Gaussian mixture models\footnote{Code and evaluation metrics more detail can be found in scikit-learn documentation (Pedregosa et al., 2011).}.
+@fig-cluster_metrics presents a detailed comparison of clustering quality across $K$-means and Gaussian mixture models^[Code and evaluation metrics detailed in @scikit-learn documentation.].
 We compute the Silhouette Score (ranging from –1 to 1; higher values indicate better cohesion and separation) to assess how closely each point is matched to its own cluster versus the next best alternative.
@@
 
-### Average Yearly Crime Rate Change by Neighbourhood Opportunity Cluster
+## Average Yearly Crime Rate Change by Neighbourhood Opportunity Cluster {#sec-tables}
 
-#### Neighbourhood Assault
+### Neighbourhood Assault
 
@@
 assault_change = assault_change.rename(
-    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+    {"Low": "Low-Opportunity ", "Med": "Medium-Opportunity", "High": "High-Opportunity"}
 )
@@
 
-#### Break-and-Enter
+### Break-and-Enter (B&E)
 
@@
 breakenter_change = breakenter_change.rename(
-    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+    {"Low": "Low-pportunity", "Med": "Medium-Opportunity", "High": "High-Opportunity"}
 )
@@
 
-#### Robbery
+### Robbery
 
@@
 robbery_change = robbery_change.rename(
-    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+    {"Low": "Low-Opportunity", "Med": "Medium-Opportunity", "High": "High-Opportunity"}
 )
@@
 
-#### Shooting
+### Shooting
 
@@
 shooting_change = shooting_change.rename(
-    {"Low": "Low Opportunity", "Med": "Medium Opportunity", "High": "High Opportunity"}
+    {"Low": "Low-Opportunity", "Med": "Medium-Opportunity", "High": "High-Opportunity"}
 )

### 2025-05-31 paper/paper.qmd
Update paper.qmd

@@
 
-We utilize the programming language Python [@python] alongside Polars [@polars] and scikit-learn [@scikit-learn] for the paper’s data cleaning and analysis.^[The workflow and GitHub-based project structure follow best practices from @tellingstories.]
+We utilize the programming language Python [@python] alongside Polars [@polars] and scikit-learn [@scikit-learn] for the paper’s data cleaning and analysis.
 
